### 2024-08-25 draft-2/draft-2.tex
initial commit

\includeruby color.rb
\includeruby syntax.rb
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\XeTeXuseglyphmetrics=1
%
\def\ldfont     #1#2#3{"#1/S=#2:mapping=tex-text#3"          at #2 pt}%
\def\ldifont    #1#2#3{"#1/S=#2/I:mapping=tex-text#3"        at #2 pt}%
\def\ldbfont    #1#2#3{"#1/S=#2/B:mapping=tex-text#3"        at #2 pt}%
\def\ldbifont   #1#2#3{"#1/S=#2/BI:mapping=tex-text#3"       at #2 pt}%
\def\ldcapfont  #1#2#3{"#1/S=#2:mapping=tex-text:+smcp#3"    at #2 pt}%
\def\ldicapfont #1#2#3{"#1/S=#2/I:mapping=tex-text:+smcp#3"  at #2 pt}%
\def\ldbcapfont #1#2#3{"#1/S=#2/B:mapping=tex-text:+smcp#3"  at #2 pt}%
\def\ldbicapfont#1#2#3{"#1/S=#2/BI:mapping=tex-text:+smcp#3" at #2 pt}%
%
\def\basefamily{Alegreya Sans}%
\def\basesize{11}%
\def\basefeat{:+liga:+clig:+calt:+onum:+pnum}%
\font\rm  =\ldfont    {\basefamily}{\basesize}{\basefeat}%
\font\it  =\ldifont   {\basefamily}{\basesize}{\basefeat}%
\font\bf  =\ldbfont   {\basefamily}{\basesize}{\basefeat}%
\font\caps=\ldcapfont {\basefamily}{\basesize}{\basefeat}%
\font\icap=\ldicapfont{\basefamily}{\basesize}{\basefeat}%
%
\def\codefamily{JetBrains Mono}%
\def\codesize{9.25}%
\def\codefeat{:+liga:+clig:+calt:lnum:+tnum}%
%
\font\fw =\ldfont {\codefamily}{\codesize}{\codefeat}%
\font\ifw=\ldifont{\codefamily}{\codesize}{\codefeat}%
%
\def\fs{%
  \spaceskip=0.5em
  \xspaceskip=0.5em
}%
%
\font\noto    ="Noto Sans:mapping=tex-text"      at 10 pt
\font\notomono="Noto Sans Mono:mapping=tex-text" at 10 pt
%
\rm
\spaceskip =0.3em plus 0.15em minus 0.15em
\xspaceskip=0.5em plus 0.25em minus 0.25em
%
\def\lcap#1{%
  \font\localfont=\ldcapfont{\basefamily}{\basesize}{\basefeat:letterspace=#1}%
  \localfont
  \dimen0=0.0625\fontdimen2\localfont
  \multiply\dimen0 by #1
  \advance\dimen0 by \fontdimen2\localfont
  \spaceskip =   \dimen0 plus 0.5\dimen0 minus 0.5\dimen0
  \xspaceskip=1.5\dimen0 plus 0.5\dimen0 minus 0.5\dimen0
}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newbox\testbox
\newcount\testem
\newcount\testwd
\def\setrpcode#1#2{%
  \setbox\testbox=\hbox{\char#2}%
  \testwd=\wd\testbox
  \multiply\testwd by 1000
  \divide\testwd by\testem
  \rpcode#1 U#2 \the\testwd
}%
\def\setlpcode#1#2{%
  \setbox\testbox=\hbox{\char#2}%
  \testwd=\wd\testbox
  \multiply\testwd by 1000
  \divide\testwd by\testem
  \lpcode#1 U#2 \the\testwd
}%
\def\setpcodes#1{%
  {#1%
    \setbox\testbox=\hbox{\kern1em}%
    \testem=\wd\testbox
    \setrpcode{#1}{"0021}% exclamation mark
    \setlpcode{#1}{"0027}% apostrophe
    \setlpcode{#1}{"0028}% left parenthesis
    \setrpcode{#1}{"0029}% right parenthesis
    \setrpcode{#1}{"002A}% asterisk
    \setrpcode{#1}{"002C}% comma
    \setrpcode{#1}{"002D}% hyphen
    \setrpcode{#1}{"002E}% full stop
    \setrpcode{#1}{"003B}% semicolon
    \setrpcode{#1}{"003A}% colon
    \setrpcode{#1}{"003F}% question mark
    \setlpcode{#1}{"005B}% left square bracket
    \setrpcode{#1}{"005D}% right square bracket
    \setlpcode{#1}{"2018}% left single quote
    \setrpcode{#1}{"2019}% right single quote
    \setlpcode{#1}{"201C}% left double quote
    \setrpcode{#1}{"201D}% right double quote
  }%
}%
\XeTeXprotrudechars=2
\setpcodes{\rm}%
\setpcodes{\it}%
\setpcodes{\bf}%
\setpcodes{\caps}%
\setpcodes{\icap}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\url#1#2{%
  \special{pdf:bann <<
    /Subtype /Link
    /A << % Action
      /S /URI % Subtype
      /URI (#1)
    >>
    /Border [ 0 0 0 ] % [ horiz corner radius, vert corner radius, width ]
    /C [ 1 0 0 ] % Color [ r, g, b ]
  >>}%
  #2\special{pdf:eann}%
}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\qt{\kern -1.0pt}%
\def\ot{\kern  0.5pt}%
\def\pt{\kern  1.0pt}%
\def\dt{\kern  2.0pt}%
\def\fem{\hskip 0.200em}%
\def\qem{\hskip 0.250em}%
\def\tem{\hskip 0.333em}%
\def\hem{\hskip 0.500em}%
\def\em {\hskip 1.000em}%
\def\dem{\hskip 2.000em}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\beginruby
  pageW   = 210
  pageH   = 297
  marginL =  50
  marginR =  50
  marginT =  38.5
  marginB =  38.5
  hoffset = (marginL - 25.4).round(3)
  voffset = (marginT - 25.4).round(3)
  hsize   = (pageW - marginL - marginR).round(3)
  vsize   = (pageH - marginT - marginB).round(3)
\endruby
\pdfpagewidth\r{pageW}mm
\pdfpageheight=\r{pageH}mm
\hoffset=\r{hoffset}mm
\voffset=\r{voffset}mm
\hsize=\r{hsize}mm
\vsize=\r{vsize}mm
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pretolerance=200
\tolerance=200
\baselineskip=16pt
\parskip=8pt
\parindent=0pt
\raggedbottom
\nopagenumbers
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\beginruby
  # Color Boxes
    grey_bg = lch(0.96, 0.0  ,   0);   grey_rbn = lch(0.70, 0.0  ,   0)
     red_bg = lch(0.96, 0.018,  15);    red_rbn = lch(0.70, 0.18 ,  15)
  yellow_bg = lch(0.97, 0.018,  90); yellow_rbn = lch(0.85, 0.15 ,  90)
   green_bg = lch(0.96, 0.018, 145);  green_rbn = lch(0.75, 0.15 , 145)
    blue_bg = lch(0.96, 0.018, 270);   blue_rbn = lch(0.70, 0.15 , 270)
  purple_bg = lch(0.96, 0.018, 330); purple_rbn = lch(0.70, 0.15 , 330)

  # Body copy
     red_fg = lch(0.6, 0.2 ,  30  )
    teal_fg = lch(0.6, 0.1 , 180  )
  purple_fg = lch(0.5, 0.25, 292.5)
    grey_fg = lch(0.5, 0.0 ,   0  )
   label_fg = lch(0.6, 0.0 ,   0  )
  ltgrey_fg = lch(0.7, 0.0 ,   0  )

  # Pseudocode
  amber_fg = lch(0.7, 0.125,  75)
  green_fg = lch(0.6, 0.15 , 150)
   blue_fg = lch(0.5, 0.2  , 270)
  lilac_fg = lch(0.6, 0.2  , 300)
\endruby
%
\newtoks\restorecolor \restorecolor={0 0 0}%
%
\long\def\color#1#2{%
  {\restorecolor={#1}\special{color rgb #1}#2}%
  \special{color rgb \the\restorecolor}%
}%
%
\long\def\colorbox#1#2#3{%
  \par
  % Set the contents of the box
  \advance\hsize by -2em
  \setbox0 = \vbox{#3}%
  \advance\hsize by 2em
  % Insert the usual interline glue
  \ifdim \prevdepth > -1000pt
    \vskip-\prevdepth
    \vskip\baselineskip
  \fi
  % Insert an extra half of a baselineskip above
  \vskip0.5\baselineskip
  % Calculate dimensions
  \dimen0 = \ht0
  \dimen1 = \dp0
  \advance\dimen0 by 0.25\baselineskip % midway along half-baselineskip above
  \advance\dimen0 by 0.25\baselineskip % adjust the upper edge and
  \advance\dimen1 by 0.20\baselineskip %   lower edge to be perceptually correct
  \advance\dimen1 by 0.25\baselineskip % midway along half-baselineskip below
  % Insert the background
  \kern -0.25\baselineskip % midway along extra
  \kern -0.25\baselineskip % adjustment
  \color{#1}{\hrule width \hsize height \dimen0 depth \dimen1}%
  \kern -\ht0
  \kern -\dp0
  \kern -0.20\baselineskip % adjustment
  \kern -0.25\baselineskip % midway along extra
  % Insert the ribbon
  \kern -0.25\baselineskip % midway along extra
  \kern -0.25\baselineskip % adjustment
  \color{#2}{\hrule width 0.5em height \dimen0 depth \dimen1}%
  \kern -\ht0
  \kern -\dp0
  \kern -0.20\baselineskip % adjustment
  \kern -0.25\baselineskip % midway along extra
  % Insert the box
  \hbox{\hskip1em\box0}
  % Insert an extra half of a baselineskip below
  \vskip0.5\baselineskip
}%
\long\def  \greybox#1{\colorbox{\r{  grey_bg}}{\r{  grey_rbn}}{#1}}%
\long\def\purplebox#1{\colorbox{\r{purple_bg}}{\r{purple_rbn}}{#1}}%
\long\def  \bluebox#1{\colorbox{\r{  blue_bg}}{\r{  blue_rbn}}{#1}}%
\long\def \greenbox#1{\colorbox{\r{ green_bg}}{\r{ green_rbn}}{#1}}%
\long\def\yellowbox#1{\colorbox{\r{yellow_bg}}{\r{yellow_rbn}}{#1}}%
\long\def   \redbox#1{\colorbox{\r{   red_bg}}{\r{   red_rbn}}{#1}}%
%
\long\def\label  #1{\color{\r{ label_fg}}{#1}}%
\long\def\term   #1{\color{\r{purple_fg}}{#1}}%
\long\def\comment#1{\color{\r{  grey_fg}}{{\noto ※ }#1}}%
\long\def\s      #1{\color{\r{  teal_fg}}{\fw #1}}%
%
\def\hex#1{{\fw\color{\r{ltgrey_fg}}{x}#1}}%
\def\uni#1{{\caps u+#1}}%
%
\def\ellipsis{.\ot.\ot.}
%
\newcount\majorcount
\newcount\minorcount
\def\major{%
  \advance\majorcount by 1
  \minorcount=0
  \leavevmode
  \llap{\label{\the\majorcount}\hem}%
}%
\def\minor{%
  \advance\minorcount by 1
  \leavevmode
  \llap{\label{\the\majorcount·\the\minorcount}\hem}%
}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\gets{<-}%
\def\to  {->}%
\def\then{=>}%
%
\setbox0=\hbox{\fw ()[]\char"7B\char"7D\it ()[]\char"7B\char"7D}%
\newdimen\strutht
\newdimen\strutdp
\strutht=\ht0
\strutdp=\dp0
\dimen0=\baselineskip
\advance\dimen0 by -\strutht
\advance\dimen0 by -\strutdp
\divide\dimen0 by 2
\advance\strutht by \dimen0
\advance\strutdp by \dimen0
\def\strut{\vrule width 0pt height \strutht depth \strutdp}%
%
\def\tab{%
  \kern 0.75ex
  \color{\r{ltgrey_fg}}{\vrule width 0.5pt}%
  \kern -0.5pt
  \kern -0.75ex
  \kern 3ex
}%
%
\def\kwd  #1{\color{\r{  blue_fg}}{#1}}% def, return, break, error
\def\flow #1{\color{\r{ amber_fg}}{#1}}% repeat, if, then, match
\def\liter#1{\color{\r{   red_fg}}{#1}}% numeric
\def\const#1{\color{\r{ lilac_fg}}{#1}}% empty, eof
\def\func #1{\color{\r{ green_fg}}{#1}}%
\def\var  #1{{\ifw #1}}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\centerline{\lcap8 UNIVERSAL CHESS INTERFACE}
\centerline{\lcap4 august 2024 • draft}

{\bf Introduction}

This specification governs the interaction between two processes named the
{\it client} and the {\it engine}. Graphical interfaces, terminal emulators,
and scripts and utilities are examples of clients. Normative text begins with
section 1 {\it Definitions}.

{\bf Getting Started}

{\it This section needs to be written! It should contain enough for a beginner
to add basic {\icap uci} support to their engine.}

{\bf Conventions}

An {\fw\color{\r{ltgrey_fg}}{x}} preceding a number indicates that it is written
in hexadecimal; for example, \hex{10} = 16. Sequences of Unicode scalar values
encoded in {\caps utf-8} are set in teal; for example, \s{u¢i} = ⟨\pt\hex{75},
\hex{c2}, \hex{a2}, \hex{69}⟩\pt.

Special terms defined by the specification are set in \term{purple} when they
are first introduced and inline comments are set \comment{after a reference
mark}.
\vskip-\parskip
\bluebox{A blue box is used to describe a convention that clients and engines
are encouraged to follow, usually to do with the interpretation or meaning of
the messages that clients and engines send. A blue box is also used to provide
a recommendation, usually to do with implementation-defined behavior.}
\vskip-\parskip
\greybox{A grey box is used to provide an explanative note or comment.}
\vskip-\parskip
\greenbox{A green box is used to provide an example of conforming behavior.}
\vskip-\parskip
Text within colored boxes is nonnormative.

\major{\bf Definitions}

\minor A \term{violation} is any violation, by the client or engine, of the
requirements of the specification. When a violation occurs, or when the
requirements of the specification are otherwise not met, the specification
imposes no further requirements on the behavior of the client or engine.

\minor An \term{error} is a condition that should not occur but that is still
governed by the specification.

\minor The engine's standard input and output must be open file descriptors and
the engine's standard error must be an open file descriptor until closed by the
engine. The sequence of bytes that the client sends to the engine via the
engine's standard input is the \term{client byte sequence}. The sequence of
bytes that the engine sends to the client via the engine's standard output
is the \term{engine byte sequence}.
\vskip-\parskip
\greybox{There are no requirements for an engine's standard error except that it
be open. For example, the engine's standard error may be directed to a null
file, to the client's standard output or standard error, or to a log file. There
are accordingly no restrictions on the sequence of bytes that the engine writes
to its standard error.}

\minor The client byte sequence and engine byte sequence must be valid
\hbox{\caps utf-8}. The sequence of scalar values encoded by the client byte
sequence is the \term{client scalar sequence} and the sequence encoded by the
engine byte sequence is the \term{engine scalar sequence}.
\vskip-\parskip
\greybox{{\caps Utf-8} is required for client−engine communication, but the
client and engine may use different encodings for other interfaces, such as
specifying file system paths to the operating system in order to read tablebase
files. For example, the client and engine may be communicating over a network
connexion and the client may be running on Linux (where paths are arbitrary
byte sequences that do not contain \hex{00} or \hex{2f}) but the engine may be
running on Windows (for which {\caps ntfs} is a common filesystem, which stores
file names in {\caps utf-16}, but where older libraries or {\caps api}s may
expect paths encoded in the local code page).}

\minor A \term{message terminator} is the pair ⟨\uni{000d carriage return},
\uni{000a line feed}⟩ or \uni{000a} alone when it is not preceded by \uni{000d}.

\minor Message terminators divide the client and engine scalar sequences into
\term{messages}; that is, a message is a (possibly empty) subsequence of scalar
values that does not contain a message terminator and every scalar value of
the client and engine scalar sequences is either part of a message or part of
a message terminator. The messages within the client sequence are \term{client
messages} and the messages within the engine sequence are \term{engine
messages}.

\vskip\parskip
\beginruby
format("let @read_byte! : File_Descriptor -> Byte | EOF")
\endruby
\vskip\parskip
\beginruby
format(%q{
def trimCR('seq : List(Byte)) : List(Byte)
  if empty?('seq) or last('seq) ≠ `0d then return 'seq
  return withoutLast('seq)
})
\endruby
\vskip\parskip
\beginruby
format(%q{
def read_message!('fd : File_Descriptor) : List(Byte) | EOF
  'seq <- empty
  repeat
    match read_byte!('fd)
      eof => return (\pt if empty?('seq) then eof else 'seq\pt)
      `0a => return trimCR('seq)
      'val => append!('seq, 'val\pt)
})
\endruby

\minor A message must not contain \uni{000d}; that is, every occurence
of \uni{000d} in the input or output sequence must be part of a message
terminator.

\minor The scalar value \uni{0020 space} divides messages into \term{tokens},
that is, a token is a non-empty sequence of scalar values that are not
\uni{0020}, and every scalar value of a message is either part of a token
or is \uni{0020}.
\vskip-\parskip
\greybox{In some cases, only the beginning of a message will be viewed as a
collection of tokens and the remainder will be viewed as a continuous sequence
of scalar values.}

\beginruby
format(%q{
def get_token!('seq : List(Byte)) -> List(Byte) | None
  repeat
    if empty?('seq) then return none
    if first('seq) ≠ `20 then break
    removeFirst!('seq)
  'tok <- empty
  repeat
    append!('tok, first('seq))
    removeFirst!('seq)
    if empty?('seq) then return 'tok
    if first('seq) = `20 then break
  repeat
    removeFirst!('seq)
    if empty?('seq) then return 'tok
    if first('seq) ≠ `20 then break
  return 'tok
})
\endruby

\vfil\break
\minor A \term{position} is a tuple with the following fields:

{\leftskip=3ex

\vskip-0.5\parskip An 8×8 array of elements, each of which is either empty
or a color−\allowbreak kind pair (where the \term{color} is white or black
and the \term{kind} is king, queen, rook, knight, bishop, or pawn). This
array is called the \term{board} and is indexed along one axis by the letters
“a” through “h” inclusive and along the other axis by the numerals “1” through
“8”. A color−\allowbreak kind pair is called a \term{piece}.

\vskip-0.5\parskip A color, white or black, called the \term{side to move}.

\vskip-0.5\parskip A collection of values, called the \term{rights}, which may
be empty or include one or more of the following: the white kingside castling
right, the white queenside castling right, the black kingside castling right,
and the black queenside castling right.

\vskip-0.5\parskip A value called the \term{en passant target}, which is either
none or one of a3, b3, \ellipsis\ot, h3, or a6, b6, \ellipsis\ot, h6.

\vskip-0.5\parskip A nonnegative integer called the \term{depth from zeroing},
which is 150 or less. \comment{This is the number of moves that have been played
(none of which are captures or pawn moves) since the last capture or pawn move,
measured in ply.}

}

\vskip-0.5\parskip The \term{side waiting} in a position is the opposite color
of the side to move.

\minor For a given position, a king is \term{in check} if it is attacked by one
or more pieces of the opposite color, where “attacked” is defined analogously to
article 3.1.2 of the 2023 {\caps fide} Laws of Chess.

\minor A position is \term{legal} if the following conditions are all satisfied:

{\leftskip=3ex

\vskip-0.5\parskip The board contains exactly one white king and one black king.

\vskip-0.5\parskip The board does not contain any pawns at indices
a1, b1, \ellipsis\ot, h1 and does not contain any pawns at indices
a8, b8, \ellipsis\ot, h8.

\vskip-0.5\parskip If the rights include the white kingside castling right, the
white king is at index e1 and there is a white rook at index h1. If the rights
include the white queenside castling right, the white king is at index e1 and
there is a white rook at index a1. If the rights include the black kingside
castling right, the black king is at index e8 and there is a black rook at
index h8. If the rights include the black queenside castling right, the black
king is at index e8 and there is a black rook at index a8.

\vskip-0.5\parskip If the en passant target is {\it n}\ot3 for some {\it n},
there is a white pawn at index {\it n}\ot4. If the en passant target is
{\it n}\ot6 for some {\it n}, there is a black pawn at index {\it n}\ot5.

\vskip-0.5\parskip The king of the color of the side waiting is not in check.

}

\vskip-\parskip
\greybox{A legal position as defined here need not be reachable from the
starting position.}

\minor A \term{move} is either a tuple with two fields – a board index
called the \term{source} and a board index called the \term{destination}
– or a tuple with three fields: a source, a destination, and a kind called
the \term{promotion kind} that is either queen, rook, bishop, or knight.

\minor For a given position {\it P}, a move {\it M} is \term{legal} and the
position {\it Q} \term{immediately follows} when {\it M} is \term{applied} if
{\it P}, {\it M}, and {\it Q} fulfill requirements analogous to articles 3.1
through 3.9 inclusive of the 2023 {\caps fide} Laws of Chess, except

{\leftskip=3ex

\vskip-0.5\parskip the requirement for a pawn to advance by two is instead that
the pawn is white and its index is one of a2, b2, \ellipsis\ot, h2 or that the
pawn is black and its index is one of a7, b7, \ellipsis\ot, h7, and

\vskip-0.5\parskip the requirement for an en passant capture is instead that
the en passant target is not none and the capturing pawn attacks the en passant
target.

}

In particular, when {\it M} is legal, there is a piece of the color of the side
to move in the board of {\it P} at the source of {\it M} and there is a piece
of the same color and kind (or a piece of the same color and the promotion kind
of {\it M}) in the board of {\it Q} at the destination of {\it M}.

The side to move of {\it Q} is the opposite color of the the side to move of
{\it P}. If {\it M} is a king or rook move, the rights of {\it Q} do not include
the corresponding castling right or castling rights. If {\it M} advances a pawn
by two, the en passant target of {\it Q} is the index between the source and
destination of {\it M}. If {\it M} is a capture or a pawn move, the depth from
zeroing of {\it Q} is zero; otherwise, the depth from zeroing of {\it Q} is the
depth from zeroing of {\it P} plus one.

\def\rjust#1#2{\hbox to #1{\hfil #2}}
\def\alt{\vrule height 7.125pt depth 1.375pt width 0.7pt}

\minor A \term{Forsyth–Edwards Notation ({\caps fen}) sequence} is a sequence
of six tokens ⟨{\it board, side to move, rights, {\icap ep} target,
{\icap dfz}, move number}⟩ of the form\par
%
\dimen0=5em
\vskip-0.5\parskip\hskip3ex\rjust{\dimen0}{\it board}
  \dt=\dt\ \r{(['{\it row}']*8).join('\pt\s/\pt')}\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it row}
  \dt=\dt\ \s8 \alt\ [\s1–\s8\s{KQRBNPkqrbnp}]+\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it side to move}
  \dt=\dt\ \s w \alt\ \s b\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it rights}
  \dt=\dt\ \s- \alt\ \s K?\pt\s Q?\pt\s k?\pt\s q?\par
  % \lower0.5pt\hbox{\notomono ∩} [\s{KQkq}]+)\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it {\icap ep} target}
  \dt=\dt\ \s- \alt\ [\s a–\s h][\s{36}]\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\icap dfz}
  \dt=\dt\ \s0 \alt\ [\s1–\s9][\s0–\s9]\lower0.25ex\hbox{*}\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it move number}
  \dt=\dt\ [\s1–\s9][\s0–\s9]\lower0.25ex\hbox{*}\par
\vskip-\parskip\hskip3ex
  \comment{The dash “\s-” is \uni{002d hyphen-minus}.}

\vskip-0.5\parskip with the following additional constraints:

{\leftskip=3ex

\vskip-0.5\parskip In each {\it row}, numerals must not be adjacent; that is,
there must not be two or more immediately consecutive numerals.

\vskip-0.5\parskip For each {\it row},
map \s K,\pt\s Q, \ellipsis\ot,\pt\s p to 1,
map \s1,\pt\s2, \ellipsis\ot,\pt\s8 to 1,\pt2,\pt.\ot.\ot.,\pt8,
and then define the width as the sum of these numbers.
For each {\it row}, the width must be 8.

\vskip-0.5\parskip The scalar values \s K and \s k must each occur exactly
once in the {\it board}.

\vskip-0.5\parskip The scalar value \s P must not occur in the first {\it row}
and must not occur in the last {\it row}, and likewise for the scalar value
\s p.

\vskip-0.5\parskip The {\icap dfz} token interpreted as a decimal integer must
be 150 or less.

\vskip-0.5\parskip The {\it move number} token interpreted as a decimal integer
must be less than 10\pt000.\par}

\minor An {\caps fen} sequence \term{describes} a position if it maps to the
position in a manner analogous to that described in article 16.1.3 of the
Portable Game Notation Specification.

\minor The \term{starting position} is the position described by the {\caps fen}
sequence\hfil\penalty-10000
\hbox{\s{rnbqkbnr/pppppppp/8/8/8/8/PPPPPPPP/RNBQKBNR w KQkq - 0 1}}.

\minor A \term{move token} is a token of the form
[\s a–\s h][\s1–\s8][\s a–\s h][\s1–\s8] or
{\it n}\ot\s7\ot{\it n}\ot\s8[\s{qrbn}] or
{\it n}\ot\s2\ot{\it n}\ot\s1[\s{qrbn}], where {\it n} is one of \s a, \s b,
\ellipsis\ot, \s h.

% \major{\bf Client Messages}

% \major{\bf Engine Messages}

% \major{\bf States and Transitions}

% \major{\bf Notation}

\bye

### 2024-08-25 draft-2/color.rb

@@ -1,0 +1,93 @@
+Tau = Math::PI * 2
+
+def linearize(x)
+  return ((x + 0.055) / 1.055)**2.4 if x > 0.04045
+  return x / 12.92
+end
+
+def delinearize(x)
+  return 1.055 * x**(1.0/2.4) - 0.055 if x > 0.0031308
+  return 12.92 * x
+end
+
+def srgb_to_oklab(r, g, b)
+  r = linearize(r)
+  g = linearize(g)
+  b = linearize(b)
+
+  l = 0.4122214708 * r + 0.5363325363 * g + 0.0514459929 * b;
+  m = 0.2119034982 * r + 0.6806995451 * g + 0.1073969566 * b;
+  s = 0.0883024619 * r + 0.2817188376 * g + 0.6299787005 * b;
+
+  l_ = Math::cbrt(l);
+  m_ = Math::cbrt(m);
+  s_ = Math::cbrt(s);
+
+  return [
+    0.2104542553 * l_ + 0.7936177850 * m_ - 0.0040720468 * s_,
+    1.9779984951 * l_ - 2.4285922050 * m_ + 0.4505937099 * s_,
+    0.0259040371 * l_ + 0.7827717662 * m_ - 0.8086757660 * s_,
+  ]
+end
+
+def oklab_to_srgb(l, a, b)
+  l_ = l + 0.396_337_777_4 * a + 0.215_803_757_3 * b
+  m_ = l - 0.105_561_345_8 * a - 0.063_854_172_8 * b
+  s_ = l - 0.089_484_177_5 * a - 1.291_485_548_0 * b
+
+  l = l_ * l_ * l_
+  m = m_ * m_ * m_
+  s = s_ * s_ * s_
+
+  return [
+    delinearize(+4.076741662_1 * l - 3.3077115913 * m + 0.2309699292 * s),
+    delinearize(-1.268438004_6 * l + 2.6097574011 * m - 0.3413193965 * s),
+    delinearize(-0.004196086_3 * l - 0.7034186147 * m + 1.7076147010 * s),
+  ]
+end
+
+def ab_to_ch(a, b)
+  c = Math::sqrt(a*a + b*b)
+  h = Math::atan2(b, a)
+  h += Tau if h < 0
+  h -= Tau if h > Tau
+  return [c, h]
+end
+
[…]

### 2024-08-25 draft-2/syntax.rb

@@ -1,0 +1,69 @@
+HILITE = {
+  'eof'       => :const,
+  'empty'     => :const,
+  'none'      => :const,
+
+  'break'     => :kwd,
+  'def'       => :kwd,
+  'error'     => :kwd,
+  'let'       => :kwd,
+  'return'    => :kwd,
+  'type'      => :kwd,
+  'violation' => :kwd,
+
+  'else'      => :flow,
+  'if'        => :flow,
+  'match'     => :flow,
+  'repeat'    => :flow,
+  'then'      => :flow,
+  'while'     => :flow,
+
+  'not'       => :func,
+  'and'       => :func,
+  'or'        => :func,
+}
+
+def format(text)
+  HILITE.each do |k, c|
+    text.gsub!(/(?<![a-zA-Z_])#{k}(?![a-zA-Z_!?])/, "\\#{c}{#{k}}")
+  end
+  # text.gsub!(/(?<![a-zA-Z0-9])([a-zA-Z_!?]+)\(/) { '\func{' + $1 + '}\pt(' }
+  # text.gsub!(/(?<=[a-zA-Z0-9])\[/, '\pt[')
+  text.gsub!(/(?<![a-zA-Z0-9])([a-zA-Z_!?]+)\(/) { '\func{' + $1 + '}(' }
+  text.gsub!(/@([a-z_!?]+)/) { '\func{' + $1 + '}' }
+  text.gsub!(/'([a-z][a-z0-9_]*)/, '\var{\1}')
+  text.gsub!(/'([0-9][0-9_]*)/, '\liter{\1}')
+  text.gsub!(/`([0-9][a-fA-F0-9_]*)/, '\liter{\hex{\1}}')
+  # text.gsub!(/( )+:( )+/, '\hskip0.5\spaceskip:\hskip0.5\spaceskip ')
+  text.gsub!(/<-( )?/) { '\gets' + (($1.nil?) ? '' : '\ ') }
+  text.gsub!(/->( )?/) { '\to'   + (($1.nil?) ? '' : '\ ') }
+  text.gsub!(/=>( )?/) { '\then' + (($1.nil?) ? '' : '\ ') }
+  # text.gsub!('...', '\dt ...\dt ')
+  # text.gsub!(?_, '\pt ')
+  text.gsub!(?_, '\hskip0.5\spaceskip ')
+  text.gsub!(?!, '{\ifw !}')
+
+  text = text.lines.map(&:rstrip)
[…]

### 2024-08-25 draft-2/draft-2.tex
insufficient condition for position legality when the en passant target is not none

@@
 \vskip-0.5\parskip If the en passant target is {\it n}\ot3 for some {\it n},
-there is a white pawn at index {\it n}\ot4. If the en passant target is
-{\it n}\ot6 for some {\it n}, there is a black pawn at index {\it n}\ot5.
+the side to move is black and there is a white pawn at index {\it n}\ot4 and at
+{\it n}\ot2 and {\it n}\ot3 the board is empty. If the en passant target is
+{\it n}\ot6 for some {\it n}, the side to move is white and there is a black
+pawn at index {\it n}\ot5 and at {\it n}\ot6 and {\it n}\ot7 the board is empty.
 

### 2024-08-25 draft-2/draft-2.tex
repunctuated the previous change to be easier to parse

--- 2024-08-25 draft-2/draft-2.tex
@@
 \vskip-0.5\parskip If the en passant target is {\it n}\ot3 for some {\it n},
-there is a white pawn at index {\it n}\ot4. If the en passant target is
-{\it n}\ot6 for some {\it n}, there is a black pawn at index {\it n}\ot5.
+then the side to move is black, there is a white pawn at index {\it n}\ot4, and
+at {\it n}\ot2 and {\it n}\ot3 the board is empty. If the en passant target is
+{\it n}\ot6 for some {\it n}, then the side to move is white, there is a black
+pawn at index {\it n}\ot5, and at {\it n}\ot6 and {\it n}\ot7 the board is
+empty.
 

### 2024-09-05 draft-2/draft-2.tex
client and engine streams are governed as sequences of bytes per se

--- 2024-08-25 draft-2/draft-2.tex
@@
 \centerline{\lcap8 UNIVERSAL CHESS INTERFACE}
-\centerline{\lcap4 august 2024 • draft}
+\centerline{\lcap4 september 2024 • draft}
 
@@
 {\it client} and the {\it engine}. Graphical interfaces, terminal emulators,
-and scripts and utilities are examples of clients. Normative text begins with
-section 1 {\it Definitions}.
+and scripts and utilities are examples of clients. Normative text begins at
+the start of section 1 {\it Definitions} and ends after section 4 {\it Engine
+Messages}.
 
@@
 
-An {\fw\color{\r{ltgrey_fg}}{x}} preceding a number indicates that it is written
-in hexadecimal; for example, \hex{10} = 16. Sequences of Unicode scalar values
-encoded in {\caps utf-8} are set in teal; for example, \s{u¢i} = ⟨\pt\hex{75},
-\hex{c2}, \hex{a2}, \hex{69}⟩\pt.
+An {\fw\color{\r{ltgrey_fg}}{x}} preceding a number indicates that it is
+written in hexadecimal; for example, \hex{10} = 16. Sequences of bytes encoded
+in {\caps ascii} are set in teal; for example, \s{uci} = ⟨\ot\hex{75}, \hex{63},
+\hex{69}⟩\pt.
 
@@
 engine. The sequence of bytes that the client sends to the engine via the
-engine's standard input is the \term{client byte sequence}. The sequence of
-bytes that the engine sends to the client via the engine's standard output
-is the \term{engine byte sequence}.
+engine's standard input is the \term{client stream}. The sequence of bytes
+that the engine sends to the client via the engine's standard output is the
+\term{engine stream}.
 \vskip-\parskip
@@
 to its standard error.}
+\vskip-\parskip
+\bluebox{No particular encoding is specified: the client and engine streams are
+not required to be encoded sequences of Unicode scalar values (or code points
+of any character set). The specification instead governs the client and engine
+streams as sequences of bytes {\it per se}.\par
+However, clients and engines are recommended to use {\caps utf-8} for
+client−engine communication so that engine, author, and option names are
+displayed properly. Note that clients and engines may use different encodings
+for other interfaces, such as for file system paths that are passed as arguments
+to the operating system. For example, the client and engine may be communicating
+over a network connexion and the client may be running on Linux (where paths
+are arbitrary byte sequences that do not contain \hex{00} or \hex{2f}) but the
+engine may be running on Windows (for which {\caps ntfs} is a common filesystem,
+which stores file names in {\caps utf-16}, but where older libraries or
+{\caps api}s may expect paths encoded in the local code page), and so the name
+of a tablebase file may require transcoding, implicitly or explicitly.}
 
-\minor The client byte sequence and engine byte sequence must be valid
-\hbox{\caps utf-8}. The sequence of scalar values encoded by the client byte
-sequence is the \term{client scalar sequence} and the sequence encoded by the
-engine byte sequence is the \term{engine scalar sequence}.
+\minor A \term{message terminator} is the pair ⟨\ot\hex{0d}, \hex{0a}⟩ or
+\hex{0a} alone when it is not preceded by \hex{0d}.
+
+\minor Message terminators divide the client and engine streams into
+\term{messages}; that is, a message is a (possibly empty) subsequence of bytes
+that does not contain a message terminator and every byte of the client and
+engine streams is either part of a message or part of a message terminator.
+The messages within the client stream are \term{client messages} and the
+messages within the engine stream are \term{engine messages}.
 \vskip-\parskip
-\greybox{{\caps Utf-8} is required for client−engine communication, but the
-client and engine may use different encodings for other interfaces, such as
-specifying file system paths to the operating system in order to read tablebase
-files. For example, the client and engine may be communicating over a network
-connexion and the client may be running on Linux (where paths are arbitrary
-byte sequences that do not contain \hex{00} or \hex{2f}) but the engine may be
-running on Windows (for which {\caps ntfs} is a common filesystem, which stores
-file names in {\caps utf-16}, but where older libraries or {\caps api}s may
-expect paths encoded in the local code page).}
+\greybox{This precludes the proper use of some encodings, such as
+{\caps utf-16} (since in {\caps utf-16}, the byte \hex{0a} appears in the
+encodings of various scalar values, whereas in {\caps utf-8}, the byte
+\hex{0a} only appears in the encoding for \uni{000a line feed}).}
 
-\minor A \term{message terminator} is the pair ⟨\uni{000d carriage return},
-\uni{000a line feed}⟩ or \uni{000a} alone when it is not preceded by \uni{000d}.
+\minor The byte \hex{20} divides messages into \term{tokens}, that is, a token
+is a non-empty sequence of bytes that are not \hex{20}, and every byte of a
+message is either part of a token or is \hex{20}.
+\vskip-\parskip
+\greybox{In some cases, only the beginning of a message will be viewed as a
+collection of tokens and the remainder will be viewed as a contiguous sequence
+of bytes.}
 
-\minor Message terminators divide the client and engine scalar sequences into
-\term{messages}; that is, a message is a (possibly empty) subsequence of scalar
-values that does not contain a message terminator and every scalar value of
-the client and engine scalar sequences is either part of a message or part of
-a message terminator. The messages within the client sequence are \term{client
-messages} and the messages within the engine sequence are \term{engine
-messages}.
+\minor A \term{position} is a tuple with the following fields:
+
+{\leftskip=3ex
+
+\vskip-0.5\parskip An 8×8 array of elements, each of which is either
+{\caps none} or a color−\allowbreak kind pair (where the \term{color} is white
+or black and the \term{kind} is king, queen, rook, bishop, knight, or pawn).
+This array is called the \term{board} and is indexed along one axis by the
+letters “a” through “h” inclusive and along the other axis by the numerals
+“1” through “8”. A color−\allowbreak kind pair is called a \term{piece}.
+
+\vskip-0.5\parskip A color, white or black, called the \term{side to move}.
+
+\vskip-0.5\parskip A collection of values, called the \term{rights}, which may
+be empty or include one or more of the following: the white kingside castling
+right, the white queenside castling right, the black kingside castling right,
+and the black queenside castling right.
+
+\vskip-0.5\parskip A value called the \term{en passant target}, which is either
+{\caps none} or one of a3, b3, \ellipsis\ot, h3, or a6, b6, \ellipsis\ot, h6.
+
+\vskip-0.5\parskip A nonnegative integer called the \term{depth from zeroing},
+which is 100 or less. \comment{This is the number of moves that have been played
+(none of which are captures or pawn moves) since the last capture or pawn move,
+measured in ply.}
+
+}
+
+\vskip-0.5\parskip The \term{side waiting} in a position is the opposite color
+of the side to move.
+
+\minor For a given position, a king is \term{in check} if it is attacked by one
+or more pieces of the opposite color, where “attacked” is defined analogously to
+article 3.1.2 of the 2023 {\caps fide} Laws of Chess.
+
+\vfil\break
+
+\minor A position is \term{valid} if the following conditions are all satisfied:
+
+{\leftskip=3ex
+
+\vskip-0.5\parskip The board contains exactly one white king and one black king.
+
+\vskip-0.5\parskip The board does not contain any pawns at indices
+a1, b1, \ellipsis\ot, h1 and does not contain any pawns at indices
+a8, b8, \ellipsis\ot, h8.
+
+\vskip-0.5\parskip If the rights include the white kingside castling right, the
+white king is at index e1 and there is a white rook at index h1. If the rights
+include the white queenside castling right, the white king is at index e1 and
+there is a white rook at index a1. If the rights include the black kingside
+castling right, the black king is at index e8 and there is a black rook at
+index h8. If the rights include the black queenside castling right, the black
+king is at index e8 and there is a black rook at index a8.
+
+\vskip-0.5\parskip If the en passant target is {\it n}\ot3 for some {\it n},
+then the side to move is black, there is a white pawn at index {\it n}\ot4,
+and at {\it n}\ot2 and {\it n}\ot3 the board contains {\caps none}. If the en
+passant target is {\it n}\ot6 for some {\it n}, then the side to move is white,
+there is a black pawn at index {\it n}\ot5, and at {\it n}\ot6 and {\it n}\ot7
+the board contains {\caps none}.
+
+\vskip-0.5\parskip The king of the color of the side waiting is not in check.
+
+\vskip-0.5\parskip There is at least one valid move that can be applied (as
+described in \the\majorcount·{\advance\minorcount by 2 \the\minorcount} below).
+\comment{This means the position is not checkmate or stalemate.}
+
+}
+
+\vskip-\parskip
+\greybox{A valid position need not be reachable from the starting position.\par
+Some engines designed for gameplay rather than analysis may additionally require
+that the positions they are sent must be reachable from the starting position.
+Such constraints are specific to engines, and are not constraints of the
+Universal Chess Interface.}
+\vskip-\parskip
+\greybox{The 2023 {\caps fide} Laws of Chess state that the game ends
+immediately when a player cannot checkmate the king by any series of legal
+moves, but this condition does not cause a position to not be valid.}
+\vskip-\parskip
+\bluebox{Engines are recommended to accept positions that are checkmate
+or stalemate even though they are not required to handle such positions.
+For such positions, engines should report the null move as the best move
+(see \the\majorcount·{\advance\minorcount by 7 \the\minorcount} below).}
+
+\minor A \term{move} is either a tuple with two fields – a board index
+called the \term{source} and a board index called the \term{destination}
+– or a tuple with three fields: a~source, a~destination, and a kind called
+the \term{promotion kind} that is either queen, rook, bishop, or knight.
+
+\minor For a given position {\it P}, a move {\it M} is \term{valid} and the
+position {\it Q} \term{immediately follows} when {\it M} is \term{applied} if
+{\it P}, {\it M}, and {\it Q} fulfill requirements analogous to articles 3.1
+through 3.9 inclusive of the 2023 {\caps fide} Laws of Chess, except
+
+{\leftskip=3ex
+
+\vskip-0.5\parskip the requirement for a pawn to advance by two is instead that
+the pawn is white and its index is one of a2, b2, \ellipsis\ot, h2 or that the
+pawn is black and its index is one of a7, b7, \ellipsis\ot, h7, and
+
+\vskip-0.5\parskip the requirement for an en passant capture is instead that
+the en passant target is not none and the capturing pawn attacks the en passant
+target.
+
+}
+
+In particular, when {\it M} is valid, there is a piece of the color of the side
+to move in the board of {\it P} at the source of {\it M} and there is a piece
+of the same color and kind (or a piece of the same color and the promotion kind
+of {\it M}) in the board of {\it Q} at the destination of {\it M}.
+
+The side to move of {\it Q} is the opposite color of the the side to move of
+{\it P}. If {\it M} is a king or rook move, the rights of {\it Q} do not include
+the corresponding castling right or castling rights. If {\it M} advances a pawn
+by two, the en passant target of {\it Q} is the index between the source and
+destination of~{\it M}. If {\it M} is a capture or a pawn move, the depth from
+zeroing of {\it Q} is zero; otherwise, the depth from zeroing of {\it Q} is the
+depth from zeroing of {\it P} plus one.
+
+\def\rjust#1#2{\hbox to #1{\hfil #2}}
+\def\alt{\vrule height 7.125pt depth 1.375pt width 0.7pt}
+
+\minor A \term{Forsyth–Edwards Notation ({\caps fen}) record} is a sequence of
+six tokens ⟨{\it board, side to move, rights, {\icap ep} target, {\icap dfz},
+move number}⟩ of the form\par
+%
+\dimen0=5em
+\vskip-0.5\parskip\hskip3ex\rjust{\dimen0}{\it board}
+  \dt=\dt\ \r{(['{\it row}']*8).join('\pt\s/\pt')}\par
+\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it row}
+  \dt=\dt\ \s8 \alt\ [\s1–\s7\s{KQRBNPkqrbnp}]+\par
+\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it side to move}
+  \dt=\dt\ \s w \alt\ \s b\par
+\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it rights}
+  \dt=\dt\ \s- \alt\ \s K?\pt\s Q?\pt\s k?\pt\s q?\par
+  % \lower0.5pt\hbox{\notomono ∩} [\s{KQkq}]+)\par
+\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it {\icap ep} target}
+  \dt=\dt\ \s- \alt\ [\s a–\s h][\s{36}]\par
+\vskip-\parskip\hskip3ex\rjust{\dimen0}{\icap dfz}
+  \dt=\dt\ \s0 \alt\ [\s1–\s9][\s0–\s9]\lower0.25ex\hbox{*}\par
+\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it move number}
+  \dt=\dt\ [\s1–\s9][\s0–\s9]\lower0.25ex\hbox{*}\par
+\vskip-\parskip\hskip3ex
+  \comment{The dash “\s-” is \hex{2d}.}
+
+\vskip-0.5\parskip with the following additional constraints:
+
+{\leftskip=3ex
+
+\vskip-0.5\parskip In each {\it row}, numerals must not be adjacent; that is,
+there must not be two or more immediately consecutive numerals.
+
+\vskip-0.5\parskip For each {\it row},
+map \s K,\pt\s Q, \ellipsis\ot,\pt\s p to 1,
+map \s1,\pt\s2, \ellipsis\ot,\pt\s8 to 1,\pt2,\pt.\ot.\ot.,\pt8,
+and then define the width as the sum of these numbers.
+For each {\it row}, the width must be 8.
+
+\vskip-0.5\parskip The bytes \s K and \s k must each occur exactly once in
+{\it board}.
+
+\vskip-0.5\parskip The byte \s P must not occur in the first {\it row} and must
+not occur in the last {\it row}, and likewise for the byte \s p.
+
+\vskip-0.5\parskip The {\icap dfz} token interpreted as a decimal integer must
+be 100 or less.
+
+\vskip-0.5\parskip The {\it move number} token interpreted as a decimal integer
+must be less than 10\pt000.\par}
+
+\minor An {\caps fen} record \term{describes} a position if it maps to the
+position in a manner analogous to that described in article 16.1.3 of the
+Portable Game Notation Specification.
+
+\minor The \term{starting position} is the position described by the {\caps fen}
+record\hfil\penalty-10000
+\hbox{\s{rnbqkbnr/pppppppp/8/8/8/8/PPPPPPPP/RNBQKBNR w KQkq - 0 1}}.
+
+\minor A \term{move token} is a token of the form
+\hbox{[\s a–\s h][\s1–\s8][\s a–\s h][\s1–\s8]} or\hfil\penalty-10000
+\hbox{[\s a–\s h]\s7[\s a–\s h]\s8[\s{qrbn}]} or
+\hbox{[\s a–\s h]\s2[\s a–\s h]\s1[\s{qrbn}]}.
+
+\minor A \term{null move token} is a token of the form \s{0000}.
+
+\major{\bf Client Messages}
+
+\major{\bf Engine Messages}
+
+\major{\bf States and Transitions}
+
+\major{\bf Notation}
+
+\vfil\break
+
+\major{\bf Examples}
 
@@
 \endruby
+
 \vskip\parskip
@@
 
-\minor A message must not contain \uni{000d}; that is, every occurence
-of \uni{000d} in the input or output sequence must be part of a message
-terminator.
-
-\minor The scalar value \uni{0020 space} divides messages into \term{tokens},
-that is, a token is a non-empty sequence of scalar values that are not
-\uni{0020}, and every scalar value of a message is either part of a token
-or is \uni{0020}.
-\vskip-\parskip
-\greybox{In some cases, only the beginning of a message will be viewed as a
-collection of tokens and the remainder will be viewed as a continuous sequence
-of scalar values.}
-
+\vskip\parskip
 \beginruby
@@
 
-\vfil\break
-\minor A \term{position} is a tuple with the following fields:
-
-{\leftskip=3ex
-
-\vskip-0.5\parskip An 8×8 array of elements, each of which is either empty
-or a color−\allowbreak kind pair (where the \term{color} is white or black
-and the \term{kind} is king, queen, rook, knight, bishop, or pawn). This
-array is called the \term{board} and is indexed along one axis by the letters
-“a” through “h” inclusive and along the other axis by the numerals “1” through
-“8”. A color−\allowbreak kind pair is called a \term{piece}.
-
-\vskip-0.5\parskip A color, white or black, called the \term{side to move}.
-
-\vskip-0.5\parskip A collection of values, called the \term{rights}, which may
-be empty or include one or more of the following: the white kingside castling
-right, the white queenside castling right, the black kingside castling right,
-and the black queenside castling right.
-
-\vskip-0.5\parskip A value called the \term{en passant target}, which is either
-none or one of a3, b3, \ellipsis\ot, h3, or a6, b6, \ellipsis\ot, h6.
-
-\vskip-0.5\parskip A nonnegative integer called the \term{depth from zeroing},
-which is 150 or less. \comment{This is the number of moves that have been played
-(none of which are captures or pawn moves) since the last capture or pawn move,
-measured in ply.}
-
-}
-
-\vskip-0.5\parskip The \term{side waiting} in a position is the opposite color
-of the side to move.
-
-\minor For a given position, a king is \term{in check} if it is attacked by one
-or more pieces of the opposite color, where “attacked” is defined analogously to
-article 3.1.2 of the 2023 {\caps fide} Laws of Chess.
-
-\minor A position is \term{legal} if the following conditions are all satisfied:
-
-{\leftskip=3ex
-
-\vskip-0.5\parskip The board contains exactly one white king and one black king.
-
-\vskip-0.5\parskip The board does not contain any pawns at indices
-a1, b1, \ellipsis\ot, h1 and does not contain any pawns at indices
-a8, b8, \ellipsis\ot, h8.
-
-\vskip-0.5\parskip If the rights include the white kingside castling right, the
-white king is at index e1 and there is a white rook at index h1. If the rights
-include the white queenside castling right, the white king is at index e1 and
-there is a white rook at index a1. If the rights include the black kingside
-castling right, the black king is at index e8 and there is a black rook at
-index h8. If the rights include the black queenside castling right, the black
-king is at index e8 and there is a black rook at index a8.
-
-\vskip-0.5\parskip If the en passant target is {\it n}\ot3 for some {\it n},
-there is a white pawn at index {\it n}\ot4. If the en passant target is
-{\it n}\ot6 for some {\it n}, there is a black pawn at index {\it n}\ot5.
-
-\vskip-0.5\parskip The king of the color of the side waiting is not in check.
-
-}
-
-\vskip-\parskip
-\greybox{A legal position as defined here need not be reachable from the
-starting position.}
-
-\minor A \term{move} is either a tuple with two fields – a board index
-called the \term{source} and a board index called the \term{destination}
-– or a tuple with three fields: a source, a destination, and a kind called
-the \term{promotion kind} that is either queen, rook, bishop, or knight.
-
-\minor For a given position {\it P}, a move {\it M} is \term{legal} and the
-position {\it Q} \term{immediately follows} when {\it M} is \term{applied} if
-{\it P}, {\it M}, and {\it Q} fulfill requirements analogous to articles 3.1
-through 3.9 inclusive of the 2023 {\caps fide} Laws of Chess, except
-
-{\leftskip=3ex
-
-\vskip-0.5\parskip the requirement for a pawn to advance by two is instead that
-the pawn is white and its index is one of a2, b2, \ellipsis\ot, h2 or that the
-pawn is black and its index is one of a7, b7, \ellipsis\ot, h7, and
-
-\vskip-0.5\parskip the requirement for an en passant capture is instead that
-the en passant target is not none and the capturing pawn attacks the en passant
-target.
-
-}
-
-In particular, when {\it M} is legal, there is a piece of the color of the side
-to move in the board of {\it P} at the source of {\it M} and there is a piece
-of the same color and kind (or a piece of the same color and the promotion kind
-of {\it M}) in the board of {\it Q} at the destination of {\it M}.
-
-The side to move of {\it Q} is the opposite color of the the side to move of
-{\it P}. If {\it M} is a king or rook move, the rights of {\it Q} do not include
-the corresponding castling right or castling rights. If {\it M} advances a pawn
-by two, the en passant target of {\it Q} is the index between the source and
-destination of {\it M}. If {\it M} is a capture or a pawn move, the depth from
-zeroing of {\it Q} is zero; otherwise, the depth from zeroing of {\it Q} is the
-depth from zeroing of {\it P} plus one.
-
-\def\rjust#1#2{\hbox to #1{\hfil #2}}
-\def\alt{\vrule height 7.125pt depth 1.375pt width 0.7pt}
-
-\minor A \term{Forsyth–Edwards Notation ({\caps fen}) sequence} is a sequence
-of six tokens ⟨{\it board, side to move, rights, {\icap ep} target,
-{\icap dfz}, move number}⟩ of the form\par
-%
-\dimen0=5em
-\vskip-0.5\parskip\hskip3ex\rjust{\dimen0}{\it board}
-  \dt=\dt\ \r{(['{\it row}']*8).join('\pt\s/\pt')}\par
-\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it row}
-  \dt=\dt\ \s8 \alt\ [\s1–\s8\s{KQRBNPkqrbnp}]+\par
-\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it side to move}
-  \dt=\dt\ \s w \alt\ \s b\par
-\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it rights}
-  \dt=\dt\ \s- \alt\ \s K?\pt\s Q?\pt\s k?\pt\s q?\par
-  % \lower0.5pt\hbox{\notomono ∩} [\s{KQkq}]+)\par
-\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it {\icap ep} target}
-  \dt=\dt\ \s- \alt\ [\s a–\s h][\s{36}]\par
-\vskip-\parskip\hskip3ex\rjust{\dimen0}{\icap dfz}
-  \dt=\dt\ \s0 \alt\ [\s1–\s9][\s0–\s9]\lower0.25ex\hbox{*}\par
-\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it move number}
-  \dt=\dt\ [\s1–\s9][\s0–\s9]\lower0.25ex\hbox{*}\par
-\vskip-\parskip\hskip3ex
-  \comment{The dash “\s-” is \uni{002d hyphen-minus}.}
-
-\vskip-0.5\parskip with the following additional constraints:
-
-{\leftskip=3ex
-
-\vskip-0.5\parskip In each {\it row}, numerals must not be adjacent; that is,
-there must not be two or more immediately consecutive numerals.
-
-\vskip-0.5\parskip For each {\it row},
-map \s K,\pt\s Q, \ellipsis\ot,\pt\s p to 1,
-map \s1,\pt\s2, \ellipsis\ot,\pt\s8 to 1,\pt2,\pt.\ot.\ot.,\pt8,
-and then define the width as the sum of these numbers.
-For each {\it row}, the width must be 8.
-
-\vskip-0.5\parskip The scalar values \s K and \s k must each occur exactly
-once in the {\it board}.
-
-\vskip-0.5\parskip The scalar value \s P must not occur in the first {\it row}
-and must not occur in the last {\it row}, and likewise for the scalar value
-\s p.
-
-\vskip-0.5\parskip The {\icap dfz} token interpreted as a decimal integer must
-be 150 or less.
-
-\vskip-0.5\parskip The {\it move number} token interpreted as a decimal integer
-must be less than 10\pt000.\par}
-
-\minor An {\caps fen} sequence \term{describes} a position if it maps to the
-position in a manner analogous to that described in article 16.1.3 of the
-Portable Game Notation Specification.
-
-\minor The \term{starting position} is the position described by the {\caps fen}
-sequence\hfil\penalty-10000
-\hbox{\s{rnbqkbnr/pppppppp/8/8/8/8/PPPPPPPP/RNBQKBNR w KQkq - 0 1}}.
-
-\minor A \term{move token} is a token of the form
-[\s a–\s h][\s1–\s8][\s a–\s h][\s1–\s8] or
-{\it n}\ot\s7\ot{\it n}\ot\s8[\s{qrbn}] or
-{\it n}\ot\s2\ot{\it n}\ot\s1[\s{qrbn}], where {\it n} is one of \s a, \s b,
-\ellipsis\ot, \s h.
-
-% \major{\bf Client Messages}
-
-% \major{\bf Engine Messages}
-
-% \major{\bf States and Transitions}
-
-% \major{\bf Notation}
-
 \bye

### 2024-09-05 draft-2/draft-2.tex
fixed typo misstating which sections are normative

@@
 This specification governs the interaction between two processes named the
-{\it client} and the {\it engine}. Graphical interfaces, terminal emulators,
-and scripts and utilities are examples of clients. Normative text begins at
-the start of section 1 {\it Definitions} and ends after section 4 {\it Engine
-Messages}.
+{\it client} and the {\it engine}. \comment{Graphical interfaces, terminal
+emulators, and scripts and utilities are examples of clients.} The text of
+sections 1–5 is normative (except as described in {\it Conventions} below).
 

### 2024-09-05 draft-2/draft-2.tex
various small changes in wording

@@
 
+\vskip3\baselineskip
+
 {\bf Introduction}
@@
 
-\minor An \term{error} is a condition that should not occur but that is still
-governed by the specification.
+% \minor An \term{error} is a condition that should not occur but that is still
+% governed by the specification.
 
@@
 not required to be encoded sequences of Unicode scalar values (or code points
-of any character set). The specification instead governs the client and engine
-streams as sequences of bytes {\it per se}.\par
+of any other character set). The specification instead governs the client and
+engine streams as sequences of bytes {\it per se}.\par
 However, clients and engines are recommended to use {\caps utf-8} for
@@
 which stores file names in {\caps utf-16}, but where older libraries or
-{\caps api}s may expect paths encoded in the local code page), and so the name
+{\caps api}s may expect paths encoded in the local code page), and so the path
 of a tablebase file may require transcoding, implicitly or explicitly.}
@@
 letters “a” through “h” inclusive and along the other axis by the numerals
-“1” through “8”. A color−\allowbreak kind pair is called a \term{piece}.
+“1” through “8” inclusive. A~color−\allowbreak kind pair is called a
+\term{piece}.
 
@@
 
-\vskip-0.5\parskip The \term{side waiting} in a position is the opposite color
+\vskip-0.5\parskip The \term{side waiting} of a position is the opposite color
 of the side to move.
@@
 then the side to move is black, there is a white pawn at index {\it n}\ot4,
-and at {\it n}\ot2 and {\it n}\ot3 the board contains {\caps none}. If the en
+and at {\it n}\ot2 and {\it n}\ot3 the board is {\caps none}. If the en
 passant target is {\it n}\ot6 for some {\it n}, then the side to move is white,
 there is a black pawn at index {\it n}\ot5, and at {\it n}\ot6 and {\it n}\ot7
-the board contains {\caps none}.
+the board is {\caps none}.
 
@@
 \vskip-\parskip
-\greybox{A valid position need not be reachable from the starting position.\par
+\greybox{A valid position (as defined above) need not be reachable from the
+starting position.\par
 Some engines designed for gameplay rather than analysis may additionally require
@@
 immediately when a player cannot checkmate the king by any series of legal
-moves, but this condition does not cause a position to not be valid.}
+moves, but this condition does not cause a position to not be valid (as defined
+above).}
 \vskip-\parskip
-\bluebox{Engines are recommended to accept positions that are checkmate
-or stalemate even though they are not required to handle such positions.
-For such positions, engines should report the null move as the best move
-(see \the\majorcount·{\advance\minorcount by 7 \the\minorcount} below).}
+\bluebox{Engines are recommended to accept positions that are checkmate or
+stalemate even though they are not required to handle receiving such positions.
+For such positions, engines should report the null move as the best move (see
+\the\majorcount·{\advance\minorcount by 7 \the\minorcount} below).}
 

### 2024-09-08 draft-2/draft-2.tex
move dfz range from definition of position to definition of valid position

@@
 \vskip-\parskip
-\greybox{This precludes the proper use of some encodings, such as
-{\caps utf-16} (since in {\caps utf-16}, the byte \hex{0a} appears in the
-encodings of various scalar values, whereas in {\caps utf-8}, the byte
-\hex{0a} only appears in the encoding for \uni{000a line feed}).}
+\greybox{This precludes the proper use of some encodings, such as {\caps utf-16}
+or {\caps utf-32} (since in {\caps utf-16} and {\caps utf-32}, the byte \hex{0a}
+appears in the encodings of various scalar values, whereas in {\caps utf-8}, the
+byte \hex{0a} only appears in the encoding for \uni{000a line feed}).}
 
@@
 of bytes.}
+
+% TODO when the remainder of a message is viewed as bytes, are both leading
+%   and trailing spaces stripped?
 
@@
 
-\vskip-0.5\parskip A nonnegative integer called the \term{depth from zeroing},
-which is 100 or less. \comment{This is the number of moves that have been played
-(none of which are captures or pawn moves) since the last capture or pawn move,
-measured in ply.}
+\vskip-0.5\parskip A nonnegative integer called the \term{depth from zeroing}.
+\comment{This is the number of moves that have been played (none of which are
+captures or pawn moves) since the last capture or pawn move, measured in ply.}
 
@@
 \vskip-0.5\parskip The king of the color of the side waiting is not in check.
+
+\vskip-0.5\parskip The depth from zeroing is 100 or less.
 

### 2024-10-02 draft-2/draft-2.tex
material for states and transitions

@@
 emulators, and scripts and utilities are examples of clients.} The text of
-sections 1–5 is normative (except as described in {\it Conventions} below).
+sections 1–4 is normative (except as described in {\it Conventions} below).
 
@@
 are arbitrary byte sequences that do not contain \hex{00} or \hex{2f}) but the
-engine may be running on Windows (for which {\caps ntfs} is a common filesystem,
+engine may be running on Windows (where {\caps ntfs} is a common filesystem,
 which stores file names in {\caps utf-16}, but where older libraries or
-{\caps api}s may expect paths encoded in the local code page), and so the path
-of a tablebase file may require transcoding, implicitly or explicitly.}
+{\caps api}s may expect paths encoded in the local code page), and so the
+path of a tablebase file may require transcoding, implicitly or explicitly.}
 
@@
 \vskip-\parskip
-\greybox{This precludes the proper use of some encodings, such as {\caps utf-16}
-or {\caps utf-32} (since in {\caps utf-16} and {\caps utf-32}, the byte \hex{0a}
-appears in the encodings of various scalar values, whereas in {\caps utf-8}, the
-byte \hex{0a} only appears in the encoding for \uni{000a line feed}).}
+\greybox{This precludes the proper use of some encodings for client−engine
+communication, such as \hbox{\caps utf-16} or \hbox{\caps utf-32} (since in
+\hbox{\caps utf-16} and \hbox{\caps utf-32}, the byte \hex{0a} appears in the
+encodings of various scalar values, whereas in \hbox{\caps utf-8}, the byte
+\hex{0a} only appears in the encoding for \uni{000a line feed}).}
+\vskip-\parskip
+\greybox{Note that \hex{00} is not disallowed.}
 
@@
 \comment{This is the number of moves that have been played (none of which are
-captures or pawn moves) since the last capture or pawn move, measured in ply.}
+captures or pawn moves) since the last capture or pawn move, measured in~ply.}
 
@@
 
+\minor {\caps todo}\hem a client message has one of seven types indicated
+by the first token of the message (\s{debug}, \s{setoption}, \s{ucinewgame},
+\s{position}, \s{isready}, \s{go}, and \s{stop})
+
 \major{\bf Engine Messages}
+
+\minor {\caps todo}\hem an engine message has one of six types indicated
+by the first token of the message (\s{id}, \s{option}, \s{uciok}, \s{info},
+\s{readyok}, \s{bestmove})
+
+\vfil\break
 
@@
 
-\major{\bf Notation}
+\minor Once the engine dequeues a \s{uci} message from the client stream,
+at every point in time the engine is in one of six \term{states}. The engine
+begins in the \term{initial} state. When the engine dequeues a client message
+or enqueues an engine message, the state may change; this is a
+\term{transition}.
+\vskip-\parskip
+\greybox{There is a particular invariant that transition rules should satisfy
+due to the nature of standard input and standard output.
+
+Firstly, the moment a message is read necessarily occurs some duration after
+the moment that the message is written (that is, the sender does not directly
+manipulate the memory of the recipient). Secondly, a process cannot both read
+from a file descriptor and write to a file descriptor as an atomic transaction
+(that is, “time of check to time of use” race conditions cannot be prevented).
+
+In particular, it is impossible for a process to guarantee that a write
+to file descriptor {\caps f} is not preceded by a write to file descriptor
+{\caps g} by another process. Even if the process attempts to read {\caps g},
+finds it empty, and immediately writes to {\caps f}, it is possible for an
+intervening write to {\caps g} to occur before the write to {\caps f} occurs.
+
+As a result, if a transition from state {\caps a} to state {\caps b} occurs
+when the engine sends a message, it is impossible to guarantee that a message
+sent by the client in state {\caps a} can actually be read by the engine in
+state {\caps a} (and not in state {\caps b}). Therefore, any message that the
+client is allowed to send in state {\caps a} should be allowed in
+state~{\caps b}. The same principle holds vice versa, and so we can state the
+invariant more generally as follows:
+
+\it For distinct processes 1 and 2, for all pairs of states {\icap a} and
+{\icap b}, if a transition from {\icap a} to {\icap b} can be caused by process
+1 sending a message, every messages that process 2 is allowed to send in state
+{\icap a} must also be allowed in state~{\icap b}.}
+
+\minor In some states, the engine may dequeue a message. If the engine dequeues
+a message, the message must be well-formed and have an allowed message type
+given the state of the engine.
+\vskip-0.5\baselineskip
+\greybox{This is not meant to imply that standard input is ever required to be
+empty, but meant to indicate that in some states the engine is not allowed to
+read standard input.}
+
+\minor In all states, the engine may enqueue a message. If the engine enqueues a
+message, the message must be well-formed and have an allowed message type given
+the state of the engine.
+
+\minor The allowed message types that may be dequeued and enqueued in each state
+are enumerated in the following table. If the {\caps next} column is not empty
+for a particular row, the action listed in the row (dequeuing or enqueuing a
+message of a particular type) causes a transition to the listed state.
+
+\dimen0=4em
+\dimen1=6em
+\dimen2=6em
+\dimen3=4em
+
+\dimen4=0em
+\advance\dimen4 by \dimen1
+\advance\dimen4 by \dimen2
+
+\def\br{\hrule height 0.8pt depth 0pt}
+\def\hr{\hrule height 0.4pt depth 0pt}
+\def\vr{\vrule width 0.4pt}
+
+\vskip\baselineskip
+\hbox to \hsize{%
+  \hfil
+  \vbox{%
+    \br
+    \hbox{\strut
+      \hbox to \dimen0{\hfil\caps state\hfil}%
+      \vr\hbox to \dimen4{\hfil\caps allowed\hfil}%
+      \vr\hbox to \dimen3{\hfil\caps next\hfil}%
+    }%
+    \br
+    \hbox{%
+      \vtop{%
+        \hbox{\strut}%
+        \kern 0.4pt\hbox to \dimen0{\strut\hfil initial\hfil}%
+      }%
+      \vr
+      \vtop{
+        \hbox{\strut}%
+        \kern 0.4pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
+      }%
+      \vtop{%
+           \hbox to \dimen2{\strut\hem\s{id}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{option}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{uciok}\hfil}%
+      }%
+      \vr
+      \vtop{%
+        \hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \hr\hbox to \dimen3{\strut\hfil idle\hfil}%
+      }%
+    }%
+    \br
+    \hbox{%
+      \vtop{%
+        \hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \kern 0.4pt
+        \kern 0.5\strutht\kern 0.5\strutdp
+        \kern 0.2pt\hbox to \dimen0{\strut\hfil idle\hfil}%
+      }%
+      \vr
+      \vtop{
+        \hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \kern 0.4pt\hbox to \dimen1{\strut\hfil\caps dequeue\hfil}%
+        \kern 0.4pt\hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \hr\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
+      }%
+      \vtop{%
+           \hbox to \dimen2{\strut\hem\s{debug}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{setoption}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{ucinewgame}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{position}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{isready}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{go}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{stop}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{info}\hfil}%
+      }%
+      \vr
+      \vtop{%
+        \hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \hr\hbox to \dimen3{\strut\hfil sync\hfil}%
+        \hr\hbox to \dimen3{\strut\hfil active\hfil}%
+        \hr\hbox{\strut}%
+      }%
+    }%
+    \br
+    \hbox{%
+      \vtop{%
+        \hbox{\strut}%
+        \kern -0.5\strutht\kern -0.5\strutdp
+        \kern 0.2pt\hbox to \dimen0{\strut\hfil sync\hfil}%
+      }%
+      \vr
+      \vtop{
+        \hbox{\strut}%
+        \kern -0.5\strutht\kern -0.5\strutdp
+        \kern 0.2pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
+      }%
+      \vtop{%
+        \hbox to \dimen2{\strut\hem\s{info}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{readyok}\hfil}%
+      }%
+      \vr
+      \vtop{%
+        \hbox{\strut}%
+        \hr\hbox to \dimen3{\strut\hfil idle\hfil}%
+      }%
+    }%
+    \br
+    \hbox{%
+      \vtop{%
+        \hbox{\strut}%
+        \kern 0.4pt\hbox{\strut}%
+        \kern 0.4pt\hbox to \dimen0{\strut\hfil active\hfil}%
+      }%
+      \vr
+      \vtop{
+        \hbox{\strut}%
+        \kern 0.4pt\hbox to \dimen1{\strut\hfil\caps dequeue\hfil}%
+        \kern 0.4pt\hbox{\strut}%
+        \hr
+        \kern 0.5\strutht\kern 0.5\strutdp
+        \kern 0.2pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
+      }%
+      \vtop{%
+           \hbox to \dimen2{\strut\hem\s{debug}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{isready}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{stop}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{info}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{bestmove}\hfil}%
+      }%
+      \vr
+      \vtop{%
+        \hbox{\strut}%
+        \hr\hbox to \dimen3{\strut\hfil ping\hfil}%
+        \hr\hbox to \dimen3{\strut\hfil halt\hfil}%
+        \hr\hbox{\strut}%
+        \hr\hbox to \dimen3{\strut\hfil idle\hfil}%
+      }%
+    }%
+    \br
+    \hbox{%
+      \vtop{%
+        \hbox{\strut}%
+        \kern -0.5\strutht\kern -0.5\strutdp
+        \kern 0.2pt\hbox to \dimen0{\strut\hfil ping\hfil}%
+      }%
+      \vr
+      \vtop{
+        \hbox{\strut}%
+        \kern -0.5\strutht\kern -0.5\strutdp
+        \kern 0.2pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
+      }%
+      \vtop{%
+        \hbox to \dimen2{\strut\hem\s{info}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{readyok}\hfil}%
+      }%
+      \vr
+      \vtop{%
+        \hbox{\strut}%
+        \hr\hbox to \dimen3{\strut\hfil active\hfil}%
+      }%
+    }%
+    \br
+    \hbox{%
+      \vtop{%
+        \hbox{\strut}%
+        \kern -0.5\strutht\kern -0.5\strutdp
+        \kern 0.2pt\hbox to \dimen0{\strut\hfil halt\hfil}%
+      }%
+      \vr
+      \vtop{
+        \hbox{\strut}%
+        \kern -0.5\strutht\kern -0.5\strutdp
+        \kern 0.2pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
+      }%
+      \vtop{%
+        \hbox to \dimen2{\strut\hem\s{info}\hfil}%
+        \hr\hbox to \dimen2{\strut\hem\s{bestmove}\hfil}%
+      }%
+      \vr
+      \vtop{%
+        \hbox{\strut}%
+        \hr\hbox to \dimen3{\strut\hfil idle\hfil}%
+      }%
+    }%
+    \br
+  }%
+  \hfil
+}
+
+{\caps todo}\hem pondering
+
+\minor {\caps todo}\hem clients are required to send \s{ucinewgame} at least
+once before the very first time they send \s{go} (but engines are strongly
+recommended to behave identically whether or not \s{ucinewgame} is ever sent
+or not)
+
+\minor {\caps todo}\hem clients are required to send \s{position} at least
+once before the very first time they send \s{go} (but engines are strongly
+recommended to behave as if \s{position startpos} were sent if \s{position}
+is not sent before \s{go} is sent)
+
+{\caps todo}\hem note that the client is not required to send \s{isready}
+before \s{go}
+
+{\caps todo}\hem note that \s{ucinewgame} does not reset the current board
+position nor does it reset the move history (only \s{position} changes the
+current board state and move history)
+
+{\caps todo}\hem note that the client may send \s{ucinewgame} after \s{position}
+before \s{go} without any additional intervening \s{position}
+
+{\caps todo}\hem note that engines and clients are not required to ignore
+or otherwise handle unknown or unexpected tokens (these are simply violations)
+
+% \major{\bf Notation}
 

### 2024-10-07 draft-2/draft-2.tex
miscellaneous cleanup and mention of void messages

@@
 \minor Message terminators divide the client and engine streams into
-\term{messages}; that is, a message is a (possibly empty) subsequence of bytes
-that does not contain a message terminator and every byte of the client and
+\term{messages}; that is, a message is a (possibly empty) sequence of bytes
+that does not contain a message terminator, and every byte of the client and
 engine streams is either part of a message or part of a message terminator.
@@
 of bytes.}
+
+\minor A message that does not contain any tokens (that is, an empty sequence
+or a sequence of one or more repetitions of \hex{20}) is \term{void}.
 
@@
 article 3.1.2 of the 2023 {\caps fide} Laws of Chess.
-
-\vfil\break
 
@@
 
-\minor {\caps todo}\hem a client message has one of seven types indicated
-by the first token of the message (\s{debug}, \s{setoption}, \s{ucinewgame},
-\s{position}, \s{isready}, \s{go}, and \s{stop})
+A client message has one of eight types indicated by the first token of the
+message (\s{uci}, \s{debug}, \s{setoption}, \s{ucinewgame}, \s{position},
+\s{isready}, \s{go}, and \s{stop}).
 
@@
 
-\minor {\caps todo}\hem an engine message has one of six types indicated
-by the first token of the message (\s{id}, \s{option}, \s{uciok}, \s{info},
-\s{readyok}, \s{bestmove})
+An engine message has one of six types indicated by the first token of the
+message (\s{id}, \s{option}, \s{uciok}, \s{info}, \s{readyok}, \s{bestmove}).
 
@@
 \major{\bf States and Transitions}
+
+\minor The client and the engine both may always enqueue or dequeue a void
+message. When the client or engine dequeues a void message, it must ignore
+the message, behaving as if the message were not transmitted.
+
+The remainder of section \the\majorcount\ is written without mention of void
+messages; consideration for void messages is tacit.
+
+% Empty messages may always be sent and must be ignored by the client
+% and engine. Empty messages never cause a transition.
 
@@
 given the state of the engine.
-\vskip-0.5\baselineskip
+\vskip-\parskip
 \greybox{This is not meant to imply that standard input is ever required to be
 empty, but meant to indicate that in some states the engine is not allowed to
-read standard input.}
+read standard input (or at least must behave as if it has not read standard
+input).}
 
@@
 are enumerated in the following table. If the {\caps next} column is not empty
-for a particular row, the action listed in the row (dequeuing or enqueuing a
-message of a particular type) causes a transition to the listed state.
+for a row, the action listed in the row (dequeuing or enqueuing a message of
+a particular type) causes a transition to the listed state.
 
@@
 }
+\vglue-3\baselineskip\vfil\break
 
-{\caps todo}\hem pondering
+\greybox{Engines and clients are not required to ignore (or otherwise handle)
+unknown or unexpected tokens (these are simply violations).}
 
-\minor {\caps todo}\hem clients are required to send \s{ucinewgame} at least
-once before the very first time they send \s{go} (but engines are strongly
-recommended to behave identically whether or not \s{ucinewgame} is ever sent
-or not)
+\minor The client must enqueue a \s{ucinewgame} message at least once before
+before it enqueues a \s{go} message for the first time.
+\vskip-\parskip
+\bluebox{Engines are strongly recommended to behave identically whether or not
+a \s{ucinewgame} message is ever sent.}
 
-\minor {\caps todo}\hem clients are required to send \s{position} at least
-once before the very first time they send \s{go} (but engines are strongly
-recommended to behave as if \s{position startpos} were sent if \s{position}
-is not sent before \s{go} is sent)
+\minor The client must enqueue a \s{position} message at least once before it
+enqueues a \s{go} message for the first time.
+\vskip-\parskip
+\bluebox{Engines are strongly recommended, if a \s{position} message is not
+sent before the first \s{go} message, to behave as if \s{position startpos}
+were sent.}
+\vskip-\parskip
+\greybox{Note that the client is not required to send \s{isready} before
+\s{go}.}
+\vskip-\parskip
+\greybox{Note that the client may send \s{ucinewgame} after \s{position}
+before \s{go} without sending any additional intervening \s{position} message.}
+\vskip-\parskip
+\greybox{Note that \s{ucinewgame} does not reset the current board state nor
+does it reset the move history. (Only \s{position} changes the current board
+state and move history.)}
 
-{\caps todo}\hem note that the client is not required to send \s{isready}
-before \s{go}
-
-{\caps todo}\hem note that \s{ucinewgame} does not reset the current board
-position nor does it reset the move history (only \s{position} changes the
-current board state and move history)
-
-{\caps todo}\hem note that the client may send \s{ucinewgame} after \s{position}
-before \s{go} without any additional intervening \s{position}
-
-{\caps todo}\hem note that engines and clients are not required to ignore
-or otherwise handle unknown or unexpected tokens (these are simply violations)
+% This specification stops governing the interaction between the client
+% and the engine when the client sends a well-formed quit message in
+% the idle state. Engines are recommended to shut down and terminate
+% themselves. Clients are recommended to provide a grace period before
+% killing the engine.
 
@@
 \major{\bf Examples}
+
+\advance\hsize by 10mm
 
@@
 format(%q{
-def trimCR('seq : List(Byte)) : List(Byte)
-  if empty?('seq) or last('seq) ≠ `0d then return 'seq
-  return withoutLast('seq)
+def withoutCR('msg : List(Byte)) : List(Byte)
+  if empty?('msg) or last('msg) ≠ `0d then return 'msg
+  return withoutLast('msg)
 })
@@
 format(%q{
-def read_message!('fd : File_Descriptor) : List(Byte) | EOF
-  'seq <- empty
+def read_message!(mut 'fd : File_Descriptor) : List(Byte) | EOF
+  'msg <- empty
   repeat
     match read_byte!('fd)
-      eof => return (\pt if empty?('seq) then eof else 'seq\pt)
-      `0a => return trimCR('seq)
-      'val => append!('seq, 'val\pt)
+      eof => return (\pt if empty?('msg) then eof else 'msg\pt)
+      `0a => return withoutCR('msg)
+      'val => append!('msg, 'val\pt)
 })
@@
 format(%q{
-def get_token!('seq : List(Byte)) -> List(Byte) | None
+def get_token!(mut 'msg : List(Byte)) : List(Byte) | None
   repeat
-    if empty?('seq) then return none
-    if first('seq) ≠ `20 then break
-    removeFirst!('seq)
+    if empty?('msg) then return none
+    if first('msg) ≠ `20 then break
+    removeFirst!('msg)
   'tok <- empty
   repeat
-    append!('tok, first('seq))
-    removeFirst!('seq)
-    if empty?('seq) then return 'tok
-    if first('seq) = `20 then break
-  repeat
-    removeFirst!('seq)
-    if empty?('seq) then return 'tok
-    if first('seq) ≠ `20 then break
+    append!('tok, first('msg))
+    removeFirst!('msg)
+    if empty?('msg) then return 'tok
+    if first('msg) = `20 then break
   return 'tok

### 2024-10-07 draft-2/syntax.rb

@@ -7,6 +7,7 @@ HILITE = {
   'def'       => :kwd,
   'error'     => :kwd,
   'let'       => :kwd,
+  'mut'       => :kwd,
   'return'    => :kwd,
   'type'      => :kwd,
   'violation' => :kwd,
@@ -38,6 +39,7 @@ def format(text)
   text.gsub!(/<-( )?/) { '\gets' + (($1.nil?) ? '' : '\ ') }
   text.gsub!(/->( )?/) { '\to'   + (($1.nil?) ? '' : '\ ') }
   text.gsub!(/=>( )?/) { '\then' + (($1.nil?) ? '' : '\ ') }
+  text.gsub!(/ +\| +/, '\hskip0.5\spaceskip|\hskip0.5\spaceskip ')
   # text.gsub!('...', '\dt ...\dt ')
   # text.gsub!(?_, '\pt ')
   text.gsub!(?_, '\hskip0.5\spaceskip ')

### 2024-10-07 draft-2/draft-2.tex
added the quit message

@@
 
-A client message has one of eight types indicated by the first token of the
+A client message has one of nine types indicated by the first token of the
 message (\s{uci}, \s{debug}, \s{setoption}, \s{ucinewgame}, \s{position},
-\s{isready}, \s{go}, and \s{stop}).
+\s{isready}, \s{go}, \s{stop}, and \s{quit}).
 
@@
 
-An engine message has one of six types indicated by the first token of the
-message (\s{id}, \s{option}, \s{uciok}, \s{info}, \s{readyok}, \s{bestmove}).
+An engine message has one of six types indicated by the first token of
+the message (\s{id}, \s{option}, \s{uciok}, \s{info}, \s{readyok}, and
+\s{bestmove}).
 
@@
 messages; consideration for void messages is tacit.
-
-% Empty messages may always be sent and must be ignored by the client
-% and engine. Empty messages never cause a transition.
 
@@
 
-% This specification stops governing the interaction between the client
-% and the engine when the client sends a well-formed quit message in
-% the idle state. Engines are recommended to shut down and terminate
-% themselves. Clients are recommended to provide a grace period before
-% killing the engine.
+\minor The specification ceases to govern the interaction between the client
+and the engine when the client enqueues a well-formed \s{quit} message.
+\vskip-\parskip
+\bluebox{Engines are recommended to stop searching immediately and exit.
+Clients are recommended, if engines are their subprocesses, to provide a
+grace period before terminating engines.}
 

### 2024-10-07 draft-2/draft-2.tex
began client messages

@@
 %
+\vglue1\baselineskip
 \centerline{\lcap8 UNIVERSAL CHESS INTERFACE}
@@
 
-\vskip3\baselineskip
+\vskip4\baselineskip
 
@@
 
-An {\fw\color{\r{ltgrey_fg}}{x}} preceding a number indicates that it is
-written in hexadecimal; for example, \hex{10} = 16. Sequences of bytes encoded
-in {\caps ascii} are set in teal; for example, \s{uci} = ⟨\ot\hex{75}, \hex{63},
-\hex{69}⟩\pt.
+Sequences are written with angle brackets; for example, ⟨\ot0, 1, 2,
+\ellipsis\pt⟩. An~{\fw\color{\r{ltgrey_fg}}{x}} preceding a number
+indicates that it is written in hexadecimal; for example, \hex{10} = 16.
+Sequences of bytes encoded in {\caps ascii} are set in teal; for example,
+\s{uci} = ⟨\ot\hex{75}, \hex{63}, \hex{69}⟩\pt. Sequences of tokens may be
+written without commas when they contain token literals; for example,
+⟨\ot\s{ab}~\s{c}~{\it d}~\s{e}\ot⟩ rather than
+⟨\ot\s{ab},~\s{c},~{\it d},~\s{e}\ot⟩.
 
@@
 
-\minor A message that does not contain any tokens (that is, an empty sequence
-or a sequence of one or more repetitions of \hex{20}) is \term{void}.
+\minor For a given sequence of tokens that begin a message (the \term{prefix}),
+the \term{suffix} is the contiguous sequence of bytes starting with the first
+byte that is not \hex{20} after the prefix and ending with the last byte of the
+message.
 
-% TODO when the remainder of a message is viewed as bytes, are both leading
-%   and trailing spaces stripped?
+\minor A sequence of bytes that does not contain any tokens (that is, an empty
+sequence or a sequence of one or more repetitions of \hex{20}) is \term{void}.
 
@@
 map \s K,\pt\s Q, \ellipsis\ot,\pt\s p to 1,
-map \s1,\pt\s2, \ellipsis\ot,\pt\s8 to 1,\pt2,\pt.\ot.\ot.,\pt8,
+map \s1,\pt\s2, \ellipsis\ot,\pt\s8 to 1,\pt2,\pt\ellipsis,\pt8,
 and then define the width as the sum of these numbers.
@@
 
-A client message has one of nine types indicated by the first token of the
-message (\s{uci}, \s{debug}, \s{setoption}, \s{ucinewgame}, \s{position},
+\minor A client message has one of nine types indicated by the first token of
+the message (\s{uci}, \s{debug}, \s{setoption}, \s{ucinewgame}, \s{position},
 \s{isready}, \s{go}, \s{stop}, and \s{quit}).
+
+\minor A \s{uci}, \s{ucinewgame}, \s{isready}, \s{stop}, or \s{quit} message
+is well-formed if it contains a single token.
+
+\minor A \s{debug} message is well-formed if it contains two tokens,
+either ⟨\s{debug on}⟩ or ⟨\s{debug off}⟩.
+\vskip-\parskip
+\bluebox{If engines have debugging information to expose via the client
+(that is, through standard output rather than standard error), they should
+feely send \s{info string} messages with this information after receiving
+\s{debug on} and suppress these messages after receiving \s{debug off}.}
+
+\minor A \s{setoption} message is well-formed if it has a prefix
+of the form ⟨\s{setoption name} \ellipsis\ \s{value}⟩, where the ellipsis is
+one or more tokens (none of which are \s{value}) matching an option name listed
+in an \s{option} message enqueued by the engine, and a suffix satisfying the
+condition(s) for the corresponding option type:
+
+{\leftskip=3ex
+
+\vskip-0.5\parskip If \s{check}, the suffix contains a single token, either
+\s{true} or \s{false}.
+
+\vskip-0.5\parskip If \s{spin}, the suffix contains a single token which is a
+decimal integer, either \s0 or of the form
+\s–?\pt[\s1–\s9][\s0–\s9]\lower0.25ex\hbox{*}, not less than the minimum value
+nor greater than the maximum value specified by the engine.
+
+\vskip-0.5\parskip If \s{combo}, the suffix contains one or more tokens and
+the sequence of tokens in the suffix is one of the sequences specified by the
+engine.
+
+\vskip-0.5\parskip If \s{string}, the suffix is not void. The suffix \s{<empty>}
+indicates an empty value. \comment{The value \s{<empty>} therefore cannot be
+specified.}
+
+}
+
+\vskip-0.5\parskip A \s{setoption} message is also well-formed if it is of the
+form ⟨\s{setoption name} \ellipsis\pt⟩ where the ellipsis is one or more tokens
+(none of which are \s{value}) matching an option name listed in an \s{option}
+message enqueued by the engine for which the corresponding option type is
+\s{button}.
+
+\minor A \s{position} message is well-formed if\ellipsis
+
+\minor A \s{go} message is well-formed if\ellipsis
 

### 2024-10-11 draft-2/draft-2.tex
added the position message

@@
 \font\icap=\ldicapfont{\basefamily}{\basesize}{\basefeat}%
+%
+\def\subfamily{Alegreya Sans Medium}%
+\def\subsize{8}%
+\font\subrm=\ldfont {\subfamily}{\subsize}{\basefeat}%
+\font\subit=\ldifont{\subfamily}{\subsize}{\basefeat}%
+%
+\def\sub#1{\lower 0.25ex\hbox{\subrm #1}}%
+\def\isub#1{\lower 0.25ex\hbox{\subit #1}}%
 %
@@
 \vskip-\parskip
-\greybox{There are no requirements for an engine's standard error except that it
-be open. For example, the engine's standard error may be directed to a null
-file, to the client's standard output or standard error, or to a log file. There
-are accordingly no restrictions on the sequence of bytes that the engine writes
-to its standard error.}
+\greybox{There are no requirements for an engine's standard error except that
+it be open. For example, the engine's standard error may be directed to a null
+device, to the client's standard output or standard error, or to a log file.
+There are accordingly no restrictions on the sequence of bytes that the engine
+writes to its standard error.}
 \vskip-\parskip
@@
 stalemate even though they are not required to handle receiving such positions.
-For such positions, engines should report the null move as the best move (see
-\the\majorcount·{\advance\minorcount by 7 \the\minorcount} below).}
-
-\minor A \term{move} is either a tuple with two fields – a board index
-called the \term{source} and a board index called the \term{destination}
-– or a tuple with three fields: a~source, a~destination, and a kind called
-the \term{promotion kind} that is either queen, rook, bishop, or knight.
+For such positions, engines should report null as the best move (see
+\the\majorcount·{\advance\minorcount by 8 \the\minorcount} below).}
+
+\minor A \term{move} is a tuple of three values: a board index called the
+\term{source}, a board index called the \term{destination}, and a field
+called the \term{promotion kind} that is either {\caps none} or queen, rook,
+bishop, or knight.
 
@@
 \vskip-\parskip\hskip3ex\rjust{\dimen0}{\icap dfz}
-  \dt=\dt\ \s0 \alt\ [\s1–\s9][\s0–\s9]\lower0.25ex\hbox{*}\par
+  \dt=\dt\ \s0 \alt\ [\s1–\s9][\s0–\s9]?\ \alt\ \s{100}\par
 \vskip-\parskip\hskip3ex\rjust{\dimen0}{\it move number}
-  \dt=\dt\ [\s1–\s9][\s0–\s9]\lower0.25ex\hbox{*}\par
+  \dt=\dt\ [\s1–\s9][\s0–\s9]\char"7B\ot0,\pt3\ot\char"7D\par
 \vskip-\parskip\hskip3ex
@@
 
-\vskip-0.5\parskip The bytes \s K and \s k must each occur exactly once in
-{\it board}.
-
-\vskip-0.5\parskip The byte \s P must not occur in the first {\it row} and must
-not occur in the last {\it row}, and likewise for the byte \s p.
-
-\vskip-0.5\parskip The {\icap dfz} token interpreted as a decimal integer must
-be 100 or less.
-
-\vskip-0.5\parskip The {\it move number} token interpreted as a decimal integer
-must be less than 10\pt000.\par}
+}
 
@@
 
-\minor A \term{move token} is a token of the form
-\hbox{[\s a–\s h][\s1–\s8][\s a–\s h][\s1–\s8]} or\hfil\penalty-10000
-\hbox{[\s a–\s h]\s7[\s a–\s h]\s8[\s{qrbn}]} or
-\hbox{[\s a–\s h]\s2[\s a–\s h]\s1[\s{qrbn}]}.
-
-\minor A \term{null move token} is a token of the form \s{0000}.
+\minor A \term{move token} is a token of the form\par
+\vskip-\parskip\hskip3ex
+  [\s a–\s h][\s1–\s8][\s a–\s h][\s1–\s8][\s{qrbn}]?\par
+\vskip-\parskip The first and second bytes of a move token interpreted in the
+natural way as a board index is the move token’s \term{source}, the third and
+fourth bytes of a move token interpreted in the natural way as a board index is
+the move token’s \term{destination}, and the fifth byte (if present) interpreted
+in the natural way as a kind is the \term{promotion kind}. If the fifth byte is
+not present, the promotion kind of the move token is {\caps none}.
+
+\minor For a given position {\it P}, a move token {\it T} \term{denotes} a
+move {\it M} if {\it M} is valid for {\it P} and the source, destination, and
+promotion kind of {\it T} match the source, destination, and promotion kind of
+{\it M} (respectively).
+
+\minor A \term{null token} is a token of the form \s{0000}.
 
@@
 \minor A \s{uci}, \s{ucinewgame}, \s{isready}, \s{stop}, or \s{quit} message
-is well-formed if it contains a single token.
-
-\minor A \s{debug} message is well-formed if it contains two tokens,
-either ⟨\s{debug on}⟩ or ⟨\s{debug off}⟩.
+is well-formed if it contains exactly one token.
+
+\minor A \s{debug} message is well-formed if it contains exactly two tokens,
+either ⟨\ot\s{debug on}\ot⟩ or ⟨\ot\s{debug off}\ot⟩.
 \vskip-\parskip
@@
 \minor A \s{setoption} message is well-formed if it has a prefix
-of the form ⟨\s{setoption name} \ellipsis\ \s{value}⟩, where the ellipsis is
+of the form ⟨\ot\s{setoption name} {\it tok}+ \s{value}\ot⟩, where {\it tok}+ is
 one or more tokens (none of which are \s{value}) matching an option name listed
@@
 
-\vskip-0.5\parskip If \s{check}, the suffix contains a single token, either
+\vskip-0.5\parskip If \s{check}, the suffix contains exactly one token, either
 \s{true} or \s{false}.
@@
 \vskip-0.5\parskip A \s{setoption} message is also well-formed if it is of the
-form ⟨\s{setoption name} \ellipsis\pt⟩ where the ellipsis is one or more tokens
+form ⟨\ot\s{setoption name} {\it tok}+\pt⟩ where {\it tok}+ is one or more tokens
 (none of which are \s{value}) matching an option name listed in an \s{option}
@@
 
-\minor A \s{position} message is well-formed if\ellipsis
+\minor A \s{position} message is well-formed if it has the form\par
+\vskip-\parskip\hskip3ex
+⟨\pt\s{position}
+  (\ot\s{startpos} \alt\ \s{fen} {\it fen}\ot⁶\pt)
+  (\ot\s{moves} {\it mov}+\ot)?\pt\ot⟩\par
+\vskip-\parskip where {\it fen}\ot⁶ is an {\caps fen} record and {\it mov}+ is
+one or more move tokens. If~\s{fen} is specified, the position described by
+{\it fen}\ot⁶ must be valid. If \s{moves} is specified:
+
+{\leftskip=3ex
+
+\vskip-0.5\parskip Let {\it P}₀ be the starting position if \s{startpos} is
+specified and the position described by {\it fen}\ot⁶ if \s{fen} is specified.
+Let {\it N} be the number of move tokens specified.
+
+\vskip-0.5\parskip For {\it n} from 0 to {\it N}−1 inclusive, the {\it n}-th
+token {\it T}\isub{n} of {\it mov}+ must denote a move {\it M}\isub{n} from
+{\it P}\isub{n}. Then let {\it P}\sub{{\subit n}+1} be the position that
+immediately follows from {\it P}\isub{n} when {\it M}\isub{n} is applied.
+
+}
 

### 2024-10-11 draft-2/draft-2.tex
the position after the last move must be valid
added superscripts

@@
 %
+\def\supfamily{Alegreya Sans Medium}%
+\def\supsize{7.4}%
+\def\supfeat{:+liga:+clig:+calt:+onum:+pnum}%
+\font\suprm=\ldfont {\supfamily}{\supsize}{\supfeat}%
+\font\supit=\ldifont{\supfamily}{\supsize}{\supfeat}%
+%
+\def\sup#1{\raise 0.73ex\hbox{\suprm #1}}%
+\def\isup#1{\raise 0.73ex\hbox{\supit #1}}%
+%
 \def\subfamily{Alegreya Sans Medium}%
-\def\subsize{8}%
-\font\subrm=\ldfont {\subfamily}{\subsize}{\basefeat}%
-\font\subit=\ldifont{\subfamily}{\subsize}{\basefeat}%
-%
-\def\sub#1{\lower 0.25ex\hbox{\subrm #1}}%
-\def\isub#1{\lower 0.25ex\hbox{\subit #1}}%
+\def\subsize{7.4}%
+\def\subfeat{:+liga:+clig:+calt:+onum:+pnum}%
+\font\subrm=\ldfont {\subfamily}{\subsize}{\subfeat}%
+\font\subit=\ldifont{\subfamily}{\subsize}{\subfeat}%
+%
+\def\sub#1{\lower 0.23ex\hbox{\subrm #1}}%
+\def\isub#1{\lower 0.23ex\hbox{\subit #1}}%
 %
@@
 
-\minor A \s{setoption} message is well-formed if it has a prefix
-of the form ⟨\ot\s{setoption name} {\it tok}+ \s{value}\ot⟩, where {\it tok}+ is
-one or more tokens (none of which are \s{value}) matching an option name listed
-in an \s{option} message enqueued by the engine, and a suffix satisfying the
-condition(s) for the corresponding option type:
+\minor A \s{setoption} message is well-formed if it has a prefix of the form
+⟨\ot\s{setoption name} {\it tok}\ot\sup+ \s{value}\ot⟩, where {\it tok}\ot\sup+
+is one or more tokens (none of which are \s{value}) matching an option name
+listed in an \s{option} message enqueued by the engine, and a suffix satisfying
+the condition(s) for the corresponding option type:
 
@@
 \vskip-0.5\parskip A \s{setoption} message is also well-formed if it is of the
-form ⟨\ot\s{setoption name} {\it tok}+\pt⟩ where {\it tok}+ is one or more tokens
-(none of which are \s{value}) matching an option name listed in an \s{option}
-message enqueued by the engine for which the corresponding option type is
-\s{button}.
+form ⟨\ot\s{setoption name} {\it tok}\ot\sup+\pt⟩ where {\it tok}\ot\sup+ is one
+or more tokens (none of which are \s{value}) matching an option name listed in
+an \s{option} message enqueued by the engine for which the corresponding option
+type is \s{button}.
 
@@
 ⟨\pt\s{position}
-  (\ot\s{startpos} \alt\ \s{fen} {\it fen}\ot⁶\pt)
-  (\ot\s{moves} {\it mov}+\ot)?\pt\ot⟩\par
-\vskip-\parskip where {\it fen}\ot⁶ is an {\caps fen} record and {\it mov}+ is
-one or more move tokens. If~\s{fen} is specified, the position described by
-{\it fen}\ot⁶ must be valid. If \s{moves} is specified:
+  (\ot\s{startpos} \alt\ \s{fen} {\it fen}\ot\sup6\pt)
+  (\ot\s{moves} {\it mov}\ot\sup+\ot)?\pt\ot⟩\par
+\vskip-\parskip where {\it fen}\ot\sup6 is an {\caps fen} record and
+{\it mov}\ot\sup+ is one or more move tokens. If~\s{fen} is specified, the
+position described by {\it fen}\ot\sup6 must be valid. Let {\it P}\sub{0} be
+the starting position if \s{startpos} is specified and the position described
+by {\it fen}\ot\sup6 if \s{fen} is specified. If \s{moves} is specified:
 
@@
 
-\vskip-0.5\parskip Let {\it P}₀ be the starting position if \s{startpos} is
-specified and the position described by {\it fen}\ot⁶ if \s{fen} is specified.
-Let {\it N} be the number of move tokens specified.
+\vskip-0.5\parskip Let {\it N} be the number of move tokens specified.
 
 \vskip-0.5\parskip For {\it n} from 0 to {\it N}−1 inclusive, the {\it n}-th
-token {\it T}\isub{n} of {\it mov}+ must denote a move {\it M}\isub{n} from
-{\it P}\isub{n}. Then let {\it P}\sub{{\subit n}+1} be the position that
+token {\it T}\isub{n} of {\it mov}\ot\sup+ must denote a move {\it M}\isub{n}
+from {\it P}\isub{n}. Then let {\it P}\sub{{\subit n}+1} be the position that
 immediately follows from {\it P}\isub{n} when {\it M}\isub{n} is applied.
 
+\vskip-0.5\parskip The position {\it P}\isub{N} must be valid.
+
 }
 
-\minor A \s{go} message is well-formed if\ellipsis
+\vskip-0.5\parskip The final position (\ot{\it P}\isub{N} if \s{moves} is
+specified, otherwise {\it P}\sub{0}\ot) is the position \term{described} by
+the \s{position} message.
+
+% TODO state machine: position message “sets” “current position” to P_N
+
+% TODO engine message: bestmove must be valid move from current position
+
+\minor A \s{go} message is well-formed if\pt\ellipsis
 

### 2024-10-12 draft-2/draft-2.tex
updated readme

@@
 \centerline{\lcap8 UNIVERSAL CHESS INTERFACE}
-\centerline{\lcap4 september 2024 • draft}
+\centerline{\lcap4 october 2024 • draft}
 
@@
 
-\minor A \term{position} is a tuple with the following fields:
+\minor A \term{position} is a structure with the following fields:
 
@@
 immediately when a player cannot checkmate the king by any series of legal
-moves, but this condition does not cause a position to not be valid (as defined
-above).}
+moves, but this condition does not cause a position to be invalid.}
 \vskip-\parskip
@@
 
-\minor A \term{move} is a tuple of three values: a board index called the
+\minor A \term{move} is a structure of three values: a board index called the
 \term{source}, a board index called the \term{destination}, and a field

### 2024-10-12 draft-2/draft-2.tex
claried articles 4·6 and 4·7

@@
 \minor The client must enqueue a \s{ucinewgame} message at least once before
-before it enqueues a \s{go} message for the first time.
+before it enqueues a \s{go} message for the first time after the initial state.
 \vskip-\parskip
@@
 \minor The client must enqueue a \s{position} message at least once before it
-enqueues a \s{go} message for the first time.
+enqueues a \s{go} message for the first time after the initial state.
 \vskip-\parskip

### 2024-10-13 draft-2/draft-2.tex
added contributors

@@
 
+% TODO remember to note in the engine section that names and values cannot
+%   contain “name” or “value”
+
 \minor A \s{setoption} message is well-formed if it has a prefix of the form
 ⟨\ot\s{setoption name} {\it tok}\ot\sup+ \s{value}\ot⟩, where {\it tok}\ot\sup+
-is one or more tokens (none of which are \s{value}) matching an option name
-listed in an \s{option} message enqueued by the engine, and a suffix satisfying
-the condition(s) for the corresponding option type:
+is one or more tokens matching an option name listed in an \s{option} message
+enqueued by the engine, and a suffix satisfying the condition(s) for the
+corresponding option type:
 
@@
 form ⟨\ot\s{setoption name} {\it tok}\ot\sup+\pt⟩ where {\it tok}\ot\sup+ is one
-or more tokens (none of which are \s{value}) matching an option name listed in
-an \s{option} message enqueued by the engine for which the corresponding option
-type is \s{button}.
+or more tokens matching an option name listed in an \s{option} message enqueued
+by the engine for which the corresponding option type is \s{button}.
 

### 2024-11-04 draft-2/draft-2.tex
added provisional description of notation

@@
 \font\ifw=\ldifont{\codefamily}{\codesize}{\codefeat}%
+%
+\font\cas=\ldfont{Cascadia Mono Italic SemiLight Italic}{9.5}{\codefeat}%
 %
@@
 %
-\def\hex#1{{\fw\color{\r{ltgrey_fg}}{x}#1}}%
+\def\hex#1{{\fw\color{\r{ltgrey_fg}}{\cas x}#1}}%
 \def\uni#1{{\caps u+#1}}%
@@
 %
-\vglue1\baselineskip
 \centerline{\lcap8 UNIVERSAL CHESS INTERFACE}
@@
 
-\vskip4\baselineskip
+\vskip3\baselineskip
 
@@
 emulators, and scripts and utilities are examples of clients.} The text of
-sections 1–4 is normative (except as described in {\it Conventions} below).
+sections 1–4 is normative (except within comments and colored boxes, described
+in the section {\it Notation} below).
 
@@
 
-{\bf Conventions}
-
-Sequences are written with angle brackets; for example, ⟨\ot0, 1, 2,
-\ellipsis\pt⟩. An~{\fw\color{\r{ltgrey_fg}}{x}} preceding a number
-indicates that it is written in hexadecimal; for example, \hex{10} = 16.
-Sequences of bytes encoded in {\caps ascii} are set in teal; for example,
-\s{uci} = ⟨\ot\hex{75}, \hex{63}, \hex{69}⟩\pt. Sequences of tokens may be
-written without commas when they contain token literals; for example,
-⟨\ot\s{ab}~\s{c}~{\it d}~\s{e}\ot⟩ rather than
-⟨\ot\s{ab},~\s{c},~{\it d},~\s{e}\ot⟩.
-
-Special terms defined by the specification are set in \term{purple} when they
-are first introduced and inline comments are set \comment{after a reference
-mark}.
+\vfil\break
+
+{\bf Notation}
+
+An~\hex{} preceding a number indicates that it is written in hexadecimal; for
+example, \hex{10} = 16. A letter or word set in teal is an abbreviation for a
+byte or sequence of bytes encoded in {\caps ascii}; for example,
+\s{k} = \hex{6b} and \s{uci} = ⟨\ot\hex{75}, \hex{63}, \hex{69}⟩\pt. Special
+terms defined by the specification are set in \term{purple} when they are
+first introduced and inline comments are set \comment{after a reference
+mark}. Comments are nonnormative.
 \vskip-\parskip
@@
 are encouraged to follow, usually to do with the interpretation or meaning of
-the messages that clients and engines send. A blue box is also used to provide
-a recommendation, usually to do with implementation-defined behavior.}
+the different messages that clients and engines can send. A blue box is also
+used to provide a recommendation, usually to do with implementation-defined
+behavior.}
 \vskip-\parskip
@@
 Text within colored boxes is nonnormative.
+
+Angle brackets are sometimes used to delimit sequences and commas sometimes used
+to separate items in a sequence. For example, ⟨\ot\s{go}~\s{depth} {\it d}
+\s{movetime}~\s{1000}\ot⟩ and
+⟨\ot\s{go},~\s{depth},~{\it d},~\s{movetime}~\s{1000}\ot⟩ are equivalent, as
+are \s{go depth 8 movetime 1000} and ⟨\ot\s{go}, \s{depth}, \s{8}, \s{movetime},
+\s{1000}\ot⟩.
+
+\def\synitem#1#2{\line{\hskip3ex\hbox to 6em{#1\hfil}\em #2\hfil}}%
+
+Patterns for sequences of bytes are written using traditional syntax for regular
+expressions:\par
+\synitem{parentheses}{group}
+\synitem{brackets}{byte class}
+\synitem{en dash}{byte range}
+\synitem{vertical bar}{alternation}
+\synitem{plus sign}{repetition (once or more)}
+%\synitem{asterisk}{repetition (zero times or more)}
+\synitem{question mark}{repetition (zero times or once)}
+\synitem{braces}{repetition ({\it n} times, or at least {\it n₁} and at most {\it n₂})}
+
+Patterns for sequences of tokens are written using analogous syntax.
+
+\vfil\break
 
@@
 \vskip-0.5\parskip\hskip3ex\rjust{\dimen0}{\it board}
-  \dt=\dt\ \r{(['{\it row}']*8).join('\pt\s/\pt')}\par
+  \dt=\dt\ {\it row}\pt(\s/\pt{\it row}\pt)\char"7B\ot7\ot\char"7D\par
 \vskip-\parskip\hskip3ex\rjust{\dimen0}{\it row}
@@
 \minor The client must enqueue a \s{ucinewgame} message at least once before
-before it enqueues a \s{go} message for the first time after the initial state.
+before it enqueues a \s{go} message for the first time after enqueuing \s{uci}.
 \vskip-\parskip
@@
 \minor The client must enqueue a \s{position} message at least once before it
-enqueues a \s{go} message for the first time after the initial state.
+enqueues a \s{go} message for the first time after enqueuing \s{uci}.
 \vskip-\parskip
@@
 \bluebox{Engines are recommended to stop searching immediately and exit.
-Clients are recommended, if engines are their subprocesses, to provide a
-grace period before terminating engines.}
-
-% \major{\bf Notation}
+Clients are recommended, if the engine is their subprocess, to provide a
+grace period before terminating the engine.}
 

### 2025-02-17 draft-2/draft-2.tex
updated month
added a small note

@@
 \centerline{\lcap8 UNIVERSAL CHESS INTERFACE}
-\centerline{\lcap4 october 2024 • draft}
+\centerline{\lcap4 november 2024 • draft}
 
@@
 
-\minor For a given sequence of tokens that begin a message (the \term{prefix}),
+\minor For a given sequence of tokens that begin a message (a \term{prefix}),
 the \term{suffix} is the contiguous sequence of bytes starting with the first
@@
 depth from zeroing of {\it P} plus one.
+
+% TODO note that this definition implies that castling moves are indicated
+%   with e1g1 rather than e1h1 (for example).
 
@@
 
+\minor A \term{search modifier} is a
+
 \minor A \s{go} message is well-formed if\pt\ellipsis

### 2025-07-24 draft-2/draft-2.tex
began work on a few items

@@
-\includeruby color.rb
-\includeruby syntax.rb
+\include color.rb
+\include syntax.rb
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\beginruby
+\begin
   pageW   = 210
@@
   vsize   = (pageH - marginT - marginB).round(3)
-\endruby
+\end
 \pdfpagewidth\r{pageW}mm
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\beginruby
+\begin
   # Color Boxes
@@
   lilac_fg = lch(0.6, 0.2  , 300)
-\endruby
+\end
 %
@@
 specified, otherwise {\it P}\sub{0}\ot) is the position \term{described} by
-the \s{position} message.
-
-% TODO state machine: position message “sets” “current position” to P_N
+the \s{position} message. The collection {\it P}\isub{0}, {\it P}\isub{1},
+\ellipsis, {\it P}\sub{{\subit N}−1} is the \term{history} of the \s{position}
+message. If \s{moves} is not specified, the history is an empty collection.
+
+\minor At any point in time, the history and position described by the
+\s{position} message most recently dequeued by the engine are the \term{current
+position history} and \term{current position}, respectively. If the engine has
+not dequeued any \s{position} messages, the current position history and current
+position are undefined. A \s{position} message is said to \term{set} the current
+position history and current position.
 
@@
 
+\minor A \term{search contextualizer} is a
+
+\minor A \term{search limit} is a
+
 \minor A \term{search modifier} is a
+
+% + 999 999 999 999 999 999 fits within  0   to +2⁶⁴−1
+% ± 999 999 999             fits within −2³¹ to +2³¹−1
+% + 9999                    fits within  0   to +2¹⁶−1
+
+% -?(0|[1−9][0−9]{0,8})   wtime btime winc binc
+%      [1−9][0−9]{0,3}    mate movestogo (cannot be zero)
+%    0|[1–9][0–9]{0,3}    depth
+%    0|[1−9][0−9]{0,8}    movetime
+%    0|[1−9][0−9]{0,17}   nodes
 
@@
 \vskip-\parskip
-\greybox{This is not meant to imply that standard input is ever required to be
-empty, but meant to indicate that in some states the engine is not allowed to
-read standard input (or at least must behave as if it has not read standard
-input).}
+\greybox{The word “may” is not meant to imply that standard input is ever
+required to be empty, but meant to indicate that in some states the engine
+is not allowed to read standard input (or at least must behave as if it has
+not read standard input).}
 
@@
 sent before the first \s{go} message, to behave as if \s{position startpos}
-were sent.}
+had been sent.}
 \vskip-\parskip
@@
 \vskip\parskip
-\beginruby
+\begin
 format("let @read_byte! : File_Descriptor -> Byte | EOF")
-\endruby
+\end
 
 \vskip\parskip
-\beginruby
+\begin
 format(%q{
@@
 })
-\endruby
+\end
 
 \vskip\parskip
-\beginruby
+\begin
 format(%q{
@@
 })
-\endruby
+\end
 
